---
title: "Imprinted Regions - Exploration through DMR plots"
author: "Gaja Matassa"
date: 'Date: `r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_document:
    toc: yes
    toc_float: yes
    theme: united
    highlight: kate
    code_folding: hide
  word_document:
    toc: yes
  pdf_document:
    toc: yes
params:
  DMRAnnotated: "~/DataDir/4.DMRAnnotated/DMRsAnnotated_final.rds"
---

# 1. Environment setting

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE)
```

```{r, collapse=TRUE}
library(ggplot2)
library(scales)
library(gridExtra)
library(dplyr)
library(dmrseq)
library(annotatr)
library(sechm)
```

# 2. Loading data

## 2.1 BSSeq objects

Loading of bsseq object.

```{r, collapse=TRUE}
bsseq_obj <- readRDS("~/DataDir/1.PreliminaryAnalysis/Output/WithSelectedEGCLCs/bsseq_obj_sharedby75ofall.rds") #getting bsseq object where sample selection and CpG filtering were already performed 
```

## 2.2 Annotated DMRs

**The DMRs here selected were called with: `delta=0.25` and `p.threshold=0.0001`**

```{r}
DMRs_annotated <- readRDS(params$DMRAnnotated)
```

## 2.3 Get imprinted regions collected from different papers

Imprinted regions were taken from the following papers:

* deSousa 2023 = https://www.science.org/doi/10.1126/sciadv.adg1936?url_ver=Z39.88-2003&rfr_id=ori:rid:crossref.org&rfr_dat=cr_pub%20%200pubmed
* Tang 2015 = https://www.cell.com/fulltext/S0092-8674%2815%2900564-4
* von Meyenn 2016 = https://www.sciencedirect.com/science/article/pii/S1534580716306372
* Court 2014 = https://genome.cshlp.org/content/24/4/554
* Kobayashi 2022 = https://www.sciencedirect.com/science/article/pii/S221367112200056X
* Jima 2022 = https://doi.org/10.1080/15592294.2022.2091815
* Babak 2015 = doi:10.1038/ng.3274

Other important references:
https://www.geneimprint.com/site/genes-by-species

* Human Imprintome Genomic Map = https://www.geneimprint.com/site/features/35786392

```{r}
ImprintedRegions <- readxl::read_xlsx("~/DataDir/5.DMRInterpretation/imprinted_region_all_for_meth.xlsx")
ImprintedRegions_GRanges <- makeGRangesFromDataFrame(ImprintedRegions, seqnames.field = "Chr", start.field = "Start", end.field = "End", keep.extra.columns = TRUE, na.rm=TRUE)
```

# 3. Get `annoTrack` with **`annotatr`** and **`dmrseq`**

```{r, collapse=TRUE}
# get annotations for hg38
annoTrack <- dmrseq::getAnnot("hg38")
```

# 4. Exploration of DMR-associated genes {.tabset}

```{r}
cellTypes_colors <- c(hiPSCs ="#dd1c77", iMeLCs ="#377eb8", hPGCLCs ="#4daf4a", hEGCLCs = "#ff7f00")
colData(bsseq_obj)$Type <- factor(colData(bsseq_obj)$Type, levels = c("hiPSCs", "iMeLCs" , "hPGCLCs", "hEGCLCs"))
```

## 4.1 hPGCLCs vs hiPSCs

### 4.1.1 Get DMRs hPGCLCs vs hiPSCs
```{r}
comparison <- "hiPSCsvshPGCLCs"
DMRs <- DMRs_annotated[[comparison]]
```

### 4.1.2 Exploration of imprinted genes within hPGCLCs vs hiPSCs DMRs

```{r, collapse=TRUE}
ImprintedGenes <- ImprintedRegions_GRanges$`Gene locus`
```

```{r, collapse=TRUE}
PresentGenes <- vector()
AbsentGenes <- vector()
for(gene_index in ImprintedGenes){
  if (any(DMRs$hgnc_symbol %in% gene_index)) {
    PresentGenes <- c(PresentGenes, gene_index)
    plotDMRs(bsseq_obj, regions=DMRs[DMRs$hgnc_symbol %in% gene_index,], testCovariate=3, annoTrack=annoTrack, qval = FALSE, stat = FALSE, extend = 5000, main = paste0(gene_index," - ", DMRs[DMRs$hgnc_symbol %in% gene_index,]$ChipSeekerAnn), verbose = FALSE, horizLegend = TRUE, col = c(rep(cellTypes_colors[1:2], each = 4), rep(cellTypes_colors[3:4], each = 3)))
  } else{ 
    AbsentGenes <- c(AbsentGenes, gene_index)
    next}
}
```

These genes were found: `r PresentGenes` These genes were not found:
`r AbsentGenes`

### 4.1.3 Exploration of irie et al., 2023 relevant genes

```{r, collapse=TRUE}
SelectedGenes <- c("CDH5", "DMRT1", "PIWIL2", "DAZL")
```

```{r, collapse=TRUE}
PresentGenes <- vector()
AbsentGenes <- vector()
for(gene_index in SelectedGenes){
  if (any(DMRs$hgnc_symbol %in% gene_index)) {
    PresentGenes <- c(PresentGenes, gene_index)
    plotDMRs(bsseq_obj, regions=DMRs[DMRs$hgnc_symbol %in% gene_index,], testCovariate=3, annoTrack=annoTrack, qval = FALSE, stat = FALSE, extend = 5000, main = paste0(gene_index," - ", DMRs[DMRs$hgnc_symbol %in% gene_index,]$ChipSeekerAnn), verbose = FALSE, horizLegend = TRUE, col = c(rep(cellTypes_colors[1:2], each = 4), rep(cellTypes_colors[3:4], each = 3)))
  } else{ 
    AbsentGenes <- c(AbsentGenes, gene_index)
    next}
}
```

These genes were found: `r PresentGenes` These genes were not found:
`r AbsentGenes`

## 4.2 hPGCLCs vs hEGCLCs

### 4.2.1 Get DMRs hPGCLCs vs hEGCLCs

```{r}
comparison <- "hEGCLCsvshPGCLCs"
DMRs <- DMRs_annotated[[comparison]]
```

### 4.2.2 Exploration of imprinted genes within hPGCLCs vs hEGCLCs DMRs

```{r, collapse=TRUE}
ImprintedGenes <- ImprintedRegions_GRanges$`Gene locus`
```

```{r, collapse=TRUE}
PresentGenes <- vector()
AbsentGenes <- vector()
for(gene_index in ImprintedGenes){
  if (any(DMRs$hgnc_symbol %in% gene_index)) {
    PresentGenes <- c(PresentGenes, gene_index)
    plotDMRs(bsseq_obj, regions=DMRs[DMRs$hgnc_symbol %in% gene_index,], testCovariate=3, annoTrack=annoTrack, qval = FALSE, stat = FALSE, extend = 5000, main = paste0(gene_index," - ", DMRs[DMRs$hgnc_symbol %in% gene_index,]$ChipSeekerAnn), verbose = FALSE, horizLegend = TRUE, col = c(rep(cellTypes_colors[1:2], each = 4), rep(cellTypes_colors[3:4], each = 3)))
  } else{ 
    AbsentGenes <- c(AbsentGenes, gene_index)
    next}
}
```

These genes were found: `r PresentGenes` These genes were not found:
`r AbsentGenes`

### 4.2.3 Exploration of irie et al., 2023 relevant genes

```{r, collapse=TRUE}
SelectedGenes <- c("CDH5", "DMRT1", "PIWIL2", "DAZL")
```

```{r, collapse=TRUE}
PresentGenes <- vector()
AbsentGenes <- vector()
for(gene_index in SelectedGenes){
  if (any(DMRs$hgnc_symbol %in% gene_index)) {
    PresentGenes <- c(PresentGenes, gene_index)
    plotDMRs(bsseq_obj, regions=DMRs[DMRs$hgnc_symbol %in% gene_index,], testCovariate=3, annoTrack=annoTrack, qval = FALSE, stat = FALSE, extend = 5000, main = paste0(gene_index," - ", DMRs[DMRs$hgnc_symbol %in% gene_index,]$ChipSeekerAnn), verbose = FALSE, horizLegend = TRUE, col = c(rep(cellTypes_colors[1:2], each = 4), rep(cellTypes_colors[3:4], each = 3)))
  } else{ 
    AbsentGenes <- c(AbsentGenes, gene_index)
    next}
}
```

These genes were found: `r PresentGenes` These genes were not found:
`r AbsentGenes`

## 4.3 hiPSCs vs hEGCLCs


### 4.3.1 Get DMRs hiPSCs vs hEGCLCs

```{r}
comparison <- "hEGCLCsvshiPSCs"
DMRs <- DMRs_annotated[[comparison]]
```

### 4.3.2 Exploration of Imprinted genes within hiPSCs vs hEGCLCs DMRs

```{r, collapse=TRUE}
ImprintedGenes <- ImprintedRegions_GRanges$`Gene locus`
```

```{r, collapse=TRUE}
PresentGenes <- vector()
AbsentGenes <- vector()
for(gene_index in ImprintedGenes){
  if (any(DMRs$hgnc_symbol %in% gene_index)) {
    PresentGenes <- c(PresentGenes, gene_index)
    plotDMRs(bsseq_obj, regions=DMRs[DMRs$hgnc_symbol %in% gene_index,], testCovariate=3, annoTrack=annoTrack, qval = FALSE, stat = FALSE, extend = 5000, main = paste0(gene_index," - ", DMRs[DMRs$hgnc_symbol %in% gene_index,]$ChipSeekerAnn), verbose = FALSE, horizLegend = TRUE, col = c(rep(cellTypes_colors[1:2], each = 4), rep(cellTypes_colors[3:4], each = 3)))
  } else{ 
    AbsentGenes <- c(AbsentGenes, gene_index)
    next}
}
```

These genes were found: `r PresentGenes` These genes were not found:
`r AbsentGenes`

### 4.3.3 Exploration of irie et al., 2023 relevant genes

```{r, collapse=TRUE}
SelectedGenes <- c("CDH5", "DMRT1", "PIWIL2", "DAZL")
```

```{r, collapse=TRUE}
PresentGenes <- vector()
AbsentGenes <- vector()
for(gene_index in SelectedGenes){
  if (any(DMRs$hgnc_symbol %in% gene_index)) {
    PresentGenes <- c(PresentGenes, gene_index)
    plotDMRs(bsseq_obj, regions=DMRs[DMRs$hgnc_symbol %in% gene_index,], testCovariate=3, annoTrack=annoTrack, qval = FALSE, stat = FALSE, extend = 5000, main = paste0(gene_index," - ", DMRs[DMRs$hgnc_symbol %in% gene_index,]$ChipSeekerAnn), verbose = FALSE, horizLegend = TRUE, col = c(rep(cellTypes_colors[1:2], each = 4), rep(cellTypes_colors[3:4], each = 3)))
  } else{ 
    AbsentGenes <- c(AbsentGenes, gene_index)
    next}
}
```

These genes were found: `r PresentGenes` These genes were not found:
`r AbsentGenes`

## 4.4  iMeLCs vs hiPSCs 

### 4.4.1 Get DMRs  iMeLCs vs hiPSCs 

```{r}
comparison <- "hiPSCsvsiMeLCs"
DMRs <- DMRs_annotated[[comparison]]
```

### 4.4.2 Exploration of imprinted genes within iMeLCs vs hiPSCs  DMRs 

```{r, collapse=TRUE}
ImprintedGenes <- ImprintedRegions_GRanges$`Gene locus`
```

```{r, collapse=TRUE}
PresentGenes <- vector()
AbsentGenes <- vector()
for(gene_index in ImprintedGenes){
  if (any(DMRs$hgnc_symbol %in% gene_index)) {
    PresentGenes <- c(PresentGenes, gene_index)
    plotDMRs(bsseq_obj, regions=DMRs[DMRs$hgnc_symbol %in% gene_index,], testCovariate=3, annoTrack=annoTrack, qval = FALSE, stat = FALSE, extend = 5000, main = paste0(gene_index," - ", DMRs[DMRs$hgnc_symbol %in% gene_index,]$ChipSeekerAnn), verbose = FALSE, horizLegend = TRUE, col = c(rep(cellTypes_colors[1:2], each = 4), rep(cellTypes_colors[3:4], each = 3)))
  } else{ 
    AbsentGenes <- c(AbsentGenes, gene_index)
    next}
}
```

These genes were found: `r PresentGenes` These genes were not found:
`r AbsentGenes`

### 4.4.3 Exploration of irie et al., 2023 relevant genes

```{r, collapse=TRUE}
SelectedGenes <- c("CDH5", "DMRT1", "PIWIL2", "DAZL")
```

```{r, collapse=TRUE}
PresentGenes <- vector()
AbsentGenes <- vector()
for(gene_index in SelectedGenes){
  if (any(DMRs$hgnc_symbol %in% gene_index)) {
    PresentGenes <- c(PresentGenes, gene_index)
    plotDMRs(bsseq_obj, regions=DMRs[DMRs$hgnc_symbol %in% gene_index,], testCovariate=3, annoTrack=annoTrack, qval = FALSE, stat = FALSE, extend = 5000, main = paste0(gene_index," - ", DMRs[DMRs$hgnc_symbol %in% gene_index,]$ChipSeekerAnn), verbose = FALSE, horizLegend = TRUE, col = c(rep(cellTypes_colors[1:2], each = 4), rep(cellTypes_colors[3:4], each = 3)))
  } else{ 
    AbsentGenes <- c(AbsentGenes, gene_index)
    next}
}
```

These genes were found: `r PresentGenes` These genes were not found:
`r AbsentGenes`

## 4.5 hPGCLCs vs iMeLCs 

### 4.5.1 Get DMRs hPGCLCs vs iMeLCs 

```{r}
comparison <- "iMeLCsvshPGCLCs"
DMRs <- DMRs_annotated[[comparison]]
```

### 4.5.2 Exploration of Imprinted genes within hPGCLCs vs iMeLCs DMRs

```{r, collapse=TRUE}
ImprintedGenes <- ImprintedRegions_GRanges$`Gene locus`
```

```{r, collapse=TRUE}
PresentGenes <- vector()
AbsentGenes <- vector()
for(gene_index in ImprintedGenes){
  if (any(DMRs$hgnc_symbol %in% gene_index)) {
    PresentGenes <- c(PresentGenes, gene_index)
    plotDMRs(bsseq_obj, regions=DMRs[DMRs$hgnc_symbol %in% gene_index,], testCovariate=3, annoTrack=annoTrack, qval = FALSE, stat = FALSE, extend = 5000, main = paste0(gene_index," - ", DMRs[DMRs$hgnc_symbol %in% gene_index,]$ChipSeekerAnn), verbose = FALSE, horizLegend = TRUE, col = c(rep(cellTypes_colors[1:2], each = 4), rep(cellTypes_colors[3:4], each = 3)))
  } else{ 
    AbsentGenes <- c(AbsentGenes, gene_index)
    next}
}
```

These genes were found: `r PresentGenes` These genes were not found:
`r AbsentGenes`

### 4.5.3 Exploration of irie et al., 2023 relevant genes

```{r, collapse=TRUE}
SelectedGenes <- c("CDH5", "DMRT1", "PIWIL2", "DAZL")
```

```{r, collapse=TRUE}
PresentGenes <- vector()
AbsentGenes <- vector()
for(gene_index in SelectedGenes){
  if (any(DMRs$hgnc_symbol %in% gene_index)) {
    PresentGenes <- c(PresentGenes, gene_index)
    plotDMRs(bsseq_obj, regions=DMRs[DMRs$hgnc_symbol %in% gene_index,], testCovariate=3, annoTrack=annoTrack, qval = FALSE, stat = FALSE, extend = 5000, main = paste0(gene_index," - ", DMRs[DMRs$hgnc_symbol %in% gene_index,]$ChipSeekerAnn), verbose = FALSE, horizLegend = TRUE, col = c(rep(cellTypes_colors[1:2], each = 4), rep(cellTypes_colors[3:4], each = 3)))
  } else{ 
    AbsentGenes <- c(AbsentGenes, gene_index)
    next}
}
```

These genes were found: `r PresentGenes` These genes were not found:
`r AbsentGenes`

## 4.6 iMeLCs vs hEGCLCs

### 4.6.1 Get DMRs  iMeLCs vs hEGCLCs

```{r}
comparison <- "hEGCLCsvsiMeLCs"
DMRs <- DMRs_annotated[[comparison]]
```

### 4.6.2 Exploration of Imprinted genes wthin iMeLCs vs hEGCLCs DMRs

```{r, collapse=TRUE}
ImprintedGenes <- ImprintedRegions_GRanges$`Gene locus`
```

```{r, collapse=TRUE}
PresentGenes <- vector()
AbsentGenes <- vector()
for(gene_index in ImprintedGenes){
  if (any(DMRs$hgnc_symbol %in% gene_index)) {
    PresentGenes <- c(PresentGenes, gene_index)
    plotDMRs(bsseq_obj, regions=DMRs[DMRs$hgnc_symbol %in% gene_index,], testCovariate=3, annoTrack=annoTrack, qval = FALSE, stat = FALSE, extend = 5000, main = paste0(gene_index," - ", DMRs[DMRs$hgnc_symbol %in% gene_index,]$ChipSeekerAnn), verbose = FALSE, horizLegend = TRUE, col = c(rep(cellTypes_colors[1:2], each = 4), rep(cellTypes_colors[3:4], each = 3)))
  } else{ 
    AbsentGenes <- c(AbsentGenes, gene_index)
    next}
}
```

These genes were found: `r PresentGenes` These genes were not found:
`r AbsentGenes`

### 4.6.3 Exploration of irie et al., 2023 relevant genes

```{r, collapse=TRUE}
SelectedGenes <- c("CDH5", "DMRT1", "PIWIL2", "DAZL")
```

```{r, collapse=TRUE}
PresentGenes <- vector()
AbsentGenes <- vector()
for(gene_index in SelectedGenes){
  if (any(DMRs$hgnc_symbol %in% gene_index)) {
    PresentGenes <- c(PresentGenes, gene_index)
    plotDMRs(bsseq_obj, regions=DMRs[DMRs$hgnc_symbol %in% gene_index,], testCovariate=3, annoTrack=annoTrack, qval = FALSE, stat = FALSE, extend = 5000, main = paste0(gene_index," - ", DMRs[DMRs$hgnc_symbol %in% gene_index,]$ChipSeekerAnn), verbose = FALSE, horizLegend = TRUE, col = c(rep(cellTypes_colors[1:2], each = 4), rep(cellTypes_colors[3:4], each = 3)))
  } else{ 
    AbsentGenes <- c(AbsentGenes, gene_index)
    next}
}
```

These genes were found: `r PresentGenes` These genes were not found:
`r AbsentGenes`


# 5. Imprinted regions in the bsseq object

```{r}
for (region in 1:length(ImprintedRegions_GRanges)) {
    tryCatch({
      plotDMRs(bsseq_obj, regions=ImprintedRegions_GRanges[region,], testCovariate=3, annoTrack=annoTrack, main = paste0("Gene locus: ", ImprintedRegions_GRanges[region,]$`Gene locus`), qval = FALSE, stat = FALSE, extend = 5000, verbose = FALSE, horizLegend = TRUE, col = c(rep(cellTypes_colors[1:2], each = 4), rep(cellTypes_colors[3:4], each = 3)))
      }, error=function(e){})
}
```

# 6. Date and Session Info

```{r}
date()
sessionInfo()
```