---
title: "DSS: Differential Methylation Analysis - Exploration of DMLtest results calling DMRs"
author: "Sarah Stucchi"
date: 'Date: `r format(Sys.Date(), "%B %d, %Y")`'
output:
    html_document:
        toc: true
        toc_float: true
        theme: 'united'
        highlight: 'kate'
        code_folding: hide
params: 
    InputFolder: '/group/testa/Project/PGCLC/Rebuttal_paperEG2025/Output_DM'
    OutputFolder: '/group/testa/Project/PGCLC/Rebuttal_paperEG2025/DM/'
---

Vignettes: 

* <https://compepigen.github.io/methrix_docs/articles/04-DMRs.html> 

* <https://bioconductor.org/packages/release/bioc/vignettes/DSS/inst/doc/DSS.html#3_Using_DSS_for_BS-seq_differential_methylation_analysis> 

* <https://compgenomr.github.io/book/bsseq.html> 


**N.B. delta=0.25, p.threshold=0.0001**

# 1. Environment setting

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE)
```

```{r, collapse=TRUE}
library(DSS)
library(plotly)
library(ggplot2)
library(scales)

source("../../Methylation_helper.R")
```

```{r, collapse=TRUE}
InputFolder <- params$InputFolder
OutputFolder <- params$OutputFolder
```

# 2. Loading of BSSeq object

Loading of bsseq object with CpGs shared by at least 75% of samples.

```{r, collapse=TRUE}
bsseq_obj <- readRDS("/group/testa/Project/PGCLC/emseq_rebuttal/2.PreliminaryAnalysis/bsseq_obj_sharedby75ofall.rds") #getting bsseq object where sample selection and CpG filtering were already performed 
bsseq_obj
```

**PCA plot**

This PCA is done on the new filtered BSseq object (different from the one in the preliminary exploration script!)

```{r}
cellTypes_colors <- c(hiPSCs ="#dd1c77", iMeLCs ="#377eb8", hPGCLCs ="#4daf4a", hEGCLCs_p5 = "#ff7f00", hEGCLCs_p10 = "#ffc400")
```

```{r, collapse=TRUE, fig.width=8}
pca_res <- do_PCA(getMeth(BSseq = bsseq_obj, type = "raw"))
plot_PCA(pca_res = pca_res, anno = pData(bsseq_obj), col_anno = "Type", shape_anno = NULL, custom_colors = cellTypes_colors, point_size = 5)
```

```{r, collapse=TRUE, fig.width=8}
plot_PCA(pca_res = pca_res, anno = pData(bsseq_obj), col_anno = "Type", shape_anno = NULL, pc_x = "PC2", pc_y = "PC3", custom_colors = cellTypes_colors, point_size = 5)
```

# 3. Differential Methylation analysis  

Loading results obtained with `DSS::DMLtest` function 

```{r, collapse=TRUE}
dml_tests_files <-  list.files(path = InputFolder, pattern = '*\\.rds', full.names = TRUE) 
#dml_tests_files <- dml_tests_files[-1]
dml_tests_files
```

Storing DMRs and DMLs
Plotting number of DMRs

```{r, collapse=TRUE}
nr_tests <- length(dml_tests_files)

DM_tests_Res <- list()
DMR_Res <- list()
DML_Res <- list()

for(i in 1:nr_tests){
  
  DM_tests_Res[[i]] <- readRDS(dml_tests_files[i])
  DMR_Res[[i]] <- list()
  DML_Res[[i]] <- list()
  
  names(DM_tests_Res)[i] <- sub(".rds", "", basename(dml_tests_files))[i]
  names(DMR_Res)[i] <- sub(".rds", "", basename(dml_tests_files))[i]
  names(DML_Res)[i] <- sub(".rds", "", basename(dml_tests_files))[i]
  
  DMR_Res[[i]] <- callDMR(DM_tests_Res[[i]], delta=0.25, p.threshold=0.0001)
  DML_Res[[i]] <- callDML(DM_tests_Res[[i]], delta=0.25, p.threshold=0.0001)
  
  }
```


```{r}
DMRbarplot(DMR=DMR_Res, comparison=1, y_lim = c(0,3300), interactive=TRUE)
```

```{r}
DMRbarplot_length(DMR=DMR_Res, comparison=1)
```

```{r}
DMRbarplot(DMR=DMR_Res, comparison=2, y_lim = c(0,3300), interactive=TRUE)
```

```{r}
DMRbarplot_length(DMR=DMR_Res, comparison=2)
```

```{r}
DMRbarplot(DMR=DMR_Res, comparison=3, y_lim = c(0,3300), interactive=TRUE)
```

```{r}
DMRbarplot_length(DMR=DMR_Res, comparison=3)
```

```{r}
DMRbarplot(DMR=DMR_Res, comparison=4, y_lim = c(0,3300), interactive=TRUE)
```

```{r}
DMRbarplot_length(DMR=DMR_Res, comparison=4)
```


**Barplot with all together**


```{r}
comparisons <- c("hEGCLCs_p10vshiPSCs", "hEGCLCs_p10vsiMeLCs", "hEGCLCs_p10vshPGCLCs", "hEGCLCs_p10vshEGCLCs_p5") #ordered with criteria

new_ylabel <- c("hEGCLCs_p10vshiPSCs" = "hiPSCs vs hEGCLCs_p10", 
                "hEGCLCs_p10vsiMeLCs" = "iMeLCs vs hEGCLCs_p10",
                "hEGCLCs_p10vshPGCLCs" = "hPGCLCs vs hEGCLCs_p10",
                "hEGCLCs_p10vshEGCLCs_p5" = "hEGCLCs_p5 vs hEGCLCs_p10")

DMRbarplot_all(DMR=DMR_Res, comparisons_ordered=comparisons, label_y=new_ylabel, y_lim = 3300, interactive=TRUE)
```



```{r}
comparisons <- c("hEGCLCs_p10vshiPSCs", "hEGCLCs_p10vshPGCLCs", "hEGCLCs_p10vshEGCLCs_p5") #ordered with criteria

new_ylabel <- c("hEGCLCs_p10vshiPSCs" = "hiPSCs vs hEGCLCs_p10", 
                "hEGCLCs_p10vshPGCLCs" = "hPGCLCs vs hEGCLCs_p10",
                "hEGCLCs_p10vshEGCLCs_p5" = "hEGCLCs_p5 vs hEGCLCs_p10")

DMRbarplot_all(DMR=DMR_Res, comparisons_ordered=comparisons, label_y=new_ylabel, y_lim = 3300, interactive=TRUE)
```



```{r}
comparisons <- c("hEGCLCs_p10vshiPSCs", "hEGCLCs_p10vshEGCLCs_p5") #ordered with criteria

new_ylabel <- c("hEGCLCs_p10vshiPSCs" = "hiPSCs vs hEGCLCs_p10", 
                "hEGCLCs_p10vshEGCLCs_p5" = "hEGCLCs_p5 vs hEGCLCs_p10")

DMRbarplot_all(DMR=DMR_Res, comparisons_ordered=comparisons, label_y=new_ylabel, y_lim = 1000, interactive=TRUE)
```



# 4. Interesting FAQ

N.B: Q: Why doesn't callDMR function return FDR for identified DMRs?

A: In DMR calling, the statistical test is perform for each CpG, and then the significant CpG are merged into regions. It is very difficult to estimate FDR at the region-level, based on the site level test results (p-values or FDR). This is a difficult question for all types of genome-wide assays such as the ChIP-seq. So I rather not to report a DMR-level FDR if it's not accurate.

# 5. Saving DMRs and DMLs and Session Info

```{r SaveRDS}
saveRDS(DM_tests_Res, paste0(OutputFolder, '/', 'DML_test_results.rds'))
saveRDS(DMR_Res, paste0(OutputFolder, '/', 'DMRs.rds'))
saveRDS(DML_Res, paste0(OutputFolder, '/', 'DMLs.rds'))
```


```{r SaveSession}
SessionInfo <- sessionInfo()
Date <- date()
```

```{r SessionInfo}
Date
SessionInfo
```
