---
title: "Generation of bsseq object with escapee CpGs (filtered with meth in hPGC > 30%) present in Hsu et al."
author: "Gaja Matassa"
date: 'Date: `r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_document:
    toc: yes
    toc_float: yes
    theme: united
    highlight: kate
    code_folding: hide
  word_document:
    toc: yes
  pdf_document:
    toc: yes
params:
  OutputFolder: '~/DataDir/7.EscapeesExploration/Output/EscapeeBsseq_Hsu/'
---

# 1. Environment setting

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, warning = FALSE)
options(scipen = 100, digits = 4)
```

```{r, collapse=TRUE}
library(ggplot2)
library(bsseq)
library(scales)
library(gridExtra)
library(dplyr)
library(dmrseq)
library(annotatr)
library(sechm)
library(rtracklayer)
source("../Methylation_helper.R")
```

```{r, collapse=TRUE}
OutputFolder <- params$OutputFolder
if (dir.exists(OutputFolder) == FALSE) {
  dir.create(OutputFolder, recursive=TRUE)
}
```

# 2. Loading data

## 2.1 BSSeq object

Loading of bsseq object with CpGs from Hsu et al.

```{r, collapse=TRUE}
bsseq_obj <- readRDS("/group/testa/Project/PGCLC/NewData_fromLaSitua/7.EscapeesExploration/bsseq_Hsu_complete.rds") 
```

## 2.3 Escapees regions

These escapees regions were filtered from the regions found in [Tang et al. 2015](https://doi.org/10.1016/j.cell.2015.04.053). In particular, those covered by at least 10 CpGs in all (Tang et al. 2015) human primordial germ cell (hPGC) samples and with methylation levels above 30% in (Tang et al. 2015) week 5-9 hPGC samples were selected.

```{r}
EscapeesRegions <- readxl::read_xlsx("~/DataDir/7.EscapeesExploration/Output/EscapeesFiltering/FirstFilteringRegions.xlsx")
```

```{r}
EscapeesRegions_GRanges_hg19 <- makeGRangesFromDataFrame(EscapeesRegions, seqnames.field = "Chr", start.field = "Start", end.field = "End", keep.extra.columns = FALSE, na.rm=TRUE)
```

We have `r length(EscapeesRegions_GRanges_hg19)` escapees regions with a total of `r sum(EscapeesRegions$CpG.counts)` CpGs.

**These regions are hg19 and need to be converted to hg38 to be compared with our data**

```{r}
path = "~/DataDir/7.EscapeesExploration/hg19ToHg38.over.chain"
ch = import.chain(path)
str(ch[[1]])

if ("UCSC" %in% seqlevelsStyle(EscapeesRegions_GRanges_hg19)) {
  EscapeesRegions_GRanges_hg38 = liftOver(EscapeesRegions_GRanges_hg19, ch)
  EscapeesRegions_GRanges_hg38 = unlist(EscapeesRegions_GRanges_hg38)  
}else{stop("The seqlevels style should be UCSC!!!")}
```


```{r}
length(EscapeesRegions_GRanges_hg38)
```

After the conversion we have `r length(EscapeesRegions_GRanges_hg38)` escapees regions.

```{r}
EscapeesRegions_GRanges <- EscapeesRegions_GRanges_hg38
```

# 3. Generation of BSSeq oject with escapee CpGs and non-escapee CpGs

```{r}
bsseq_ranges_df <- granges(bsseq_obj) %>% as.data.frame() #granges has 0-based position they seem 1-based. Look here for explanation: https://www.biostars.org/p/84686/
bsseq_ranges_df$end <- bsseq_ranges_df$start + 1 #to make them 0-based
bsseq_granges <- makeGRangesFromDataFrame(bsseq_ranges_df, starts.in.df.are.0based = TRUE)

bsseq_obj_1based <- bsseq_obj
bsseq_obj_1based@rowRanges <- bsseq_granges
granges(bsseq_obj_1based)
```

This to obtain a bsseq object with only CpGs contained in the escapees regions

```{r}
m <- findOverlaps(bsseq_granges, EscapeesRegions_GRanges)

bsseq_obj_escapees <- bsseq_obj_1based[queryHits(m),]
```

With our targeted kit we are able to cover **`r length(bsseq_obj_escapees)`** CpGs belonging to all the escapees regions (`r round((length(bsseq_obj_escapees)/sum(EscapeesRegions$CpG.counts))*100, 2)`%).

Let's create also a bsseq object with non-escapee CpGs (all the other CpGs that are not in escapees regions)

```{r}
bsseq_obj_NonEscapees <- bsseq_obj_1based[-queryHits(m),]
```

# 4. Violin plots

## 4.1 showing global trend (and mean) of 5mCs + 5hmCs in escapee regions

```{r}
violin_plot(assays(bsseq_obj_escapees)$Meth_WGBS, stat="median")
```

## 4.2 showing global trend (and mean) of only 5hmCs in escapee regions

```{r}
violin_plot(assays(bsseq_obj_escapees)$Meth_hmC, stat="median")
```

## 4.3 showing global trend (and mean) of only 5mCs in escapee regions

```{r}
violin_plot(assays(bsseq_obj_escapees)$Meth_mC, stat="median")
```

# 5. Saving and Session Info

Saving:

-   DataFrame of Escapee regions after filtering and conversion to hg38 (excel format)

-   bsseq object subset with only CpGs in escapees regions where hPGCs have meth \> 0.3

-   bsseq object subset with non-escapee CpGs (all the other CpGs that are not in escapees regions)

```{r}
#openxlsx::write.xlsx(as.data.frame(EscapeesRegions_GRanges) %>% rename(chr = seqnames), file = paste0(OutputFolder, "EscapeeRegions_hg38.xlsx")) 
saveRDS(bsseq_obj_escapees, paste0(OutputFolder, "bsseq_obj_escapees_Hsu.rds"))
saveRDS(bsseq_obj_NonEscapees, paste0(OutputFolder, "bsseq_obj_NonEscapees_Hsu.rds"))
```

```{r}
date()
sessionInfo()
```
