---
title: "Heatmaps of escapees CpG methylation levels"
author: "Gaja Matassa"
date: 'Date: `r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_document:
    toc: yes
    toc_float: yes
    theme: united
    highlight: kate
    code_folding: hide
  word_document:
    toc: yes
  pdf_document:
    toc: yes
params:
  InputFolder: '~/DataDir/7.EscapeesExploration/Output/EscapeeBsseq_Hsu/'
---

# 1. Environment setting

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, warning = FALSE)
options(scipen = 100, digits = 4)
```

```{r, collapse=TRUE}
library(dplyr)
library(ggplot2)
library(scales)
library(gridExtra)
library(bsseq)
library(sechm)
    
source("../Methylation_helper.R")
```

```{r, collapse=TRUE}
InputFolder <- params$InputFolder
```

# 2. Loading data

Loading bsseq object
```{r}
bsseq_obj_escapees <- readRDS(paste0(InputFolder, "bsseq_obj_escapees_Hsu.rds"))
```

Loading annotated escapees
```{r}
AnnotatedEscapees <- readRDS("~/DataDir/7.EscapeesExploration/Output/EscapeesAnnotationAndFunctional_Hsu/AnnotatedEscapees.rds")
```

# 3. Heatmap

CpGs already filtered to be covered by ALL samples

We have `r length(bsseq_obj_escapees)` CpGs.

Add annotation made in script 2.EscapeeAnnotationAndFunctional.Rmd

```{r}
AnnotatedEscapees_gr <- makeGRangesFromDataFrame(AnnotatedEscapees, keep.extra.columns = TRUE)
```

```{r}
m <- findOverlaps(rowRanges(bsseq_obj_escapees), AnnotatedEscapees_gr)

bsseq_obj_subset <- bsseq_obj_escapees[queryHits(m),]

rowData(bsseq_obj_subset)$ensembl_gene_id <- AnnotatedEscapees_gr[subjectHits(m), ]$ensembl_gene_id
```

Let's show the __average__ methylation levels. Instead of plotting for each row an escapee CpGs, I aggregate CpGs by ensembl gene id __averaging__ the methylation levels. 

## 3.1 5mCs + 5hmCs (WGBS)

```{r, fig.width=12, fig.height=15}
AvgMeth <- aggregate(assays(bsseq_obj_subset)$Meth_WGBS, list(rowData(bsseq_obj_subset)$ensembl_gene_id), mean, na.rm = T)
rownames(AvgMeth) <- AvgMeth$Group.1
AvgMeth <- AvgMeth[, -1]

rowDataSE <- rowData(bsseq_obj_subset)
rownames(rowDataSE) <- rowData(bsseq_obj_subset)$ensembl_gene_id
rowDataSE <- rowDataSE[rownames(AvgMeth), ]

SampleAnno <- colData(bsseq_obj_subset)

SE <- SummarizedExperiment(assays=AvgMeth, colData=SampleAnno, rowData = rowDataSE)

identical(colnames(assay(SE)), rownames(colData(SE))) #TRUE

# cellTypes_colors <- c(hiPSCs ="#dd1c77", iMeLCs ="#377eb8", hPGCLCs ="#4daf4a", hEGCLCs = "#ff7f00")
# colData(SE)$Type <- factor(colData(SE)$Type, levels = c("hiPSCs", "iMeLCs" , "hPGCLCs", "hEGCLCs"))
# 
# metadata(SE)$anno_colors <- list(Type = cellTypes_colors,
#                                                GenomicRegion = c ("3' UTR" = "#FFADAD", "5' UTR" = "#FFD6A5", "Distal Intergenic" = "#FDFFB6", "Downstream (<=300bp)" = "#FFFFFC", "Exon" = "#9BF6FF", "Intron" = "#A0C4FF", "Promoter (<=1kb)" = "#BDB2FF", "Promoter (1-2kb)" = "#FFC6FF", "Promoter (2-3kb)" = "#CAFFBF"))
```

Number of rows is `r length(AvgMeth)` therefore too many for an heatmap. I'll take only the first 3000.

```{r}
break_coord <- assay(SE) %>% quantile(seq(0,1, by=0.1), na.rm =TRUE)
ExpCols <- viridisLite::plasma(length(seq(head(break_coord,1), tail(break_coord,1), by=0.2)))
sechm(SE, features = rownames(SE)[1:3000], hmcols=ExpCols, breaks=seq(head(break_coord,1), tail(break_coord,1), by=0.2), show_rownames=FALSE, show_colnames=TRUE, do.scale=FALSE, column_title = "Average methylation levels (by gene) of CpGs in escapee regions", top_annotation = c("Type"), row_title_rot = 0, cluster_rows = TRUE, gaps_at = "Type")
```

Let's change the range of methylation

```{r}
break_coord <- assay(SE) %>% quantile(seq(0,1, by=0.1), na.rm =TRUE)
ExpCols <- viridisLite::plasma(length(seq(break_coord[2], tail(break_coord,1), by=0.05)))
sechm(SE, features = rownames(SE)[1:3000], hmcols=ExpCols, breaks=seq(break_coord[2], tail(break_coord,1), by=0.05), show_rownames=FALSE, show_colnames=TRUE, do.scale=FALSE, column_title = "Average methylation levels (by gene) of CpGs in escapee regions", top_annotation = c("Type"), row_title_rot = 0, cluster_rows = TRUE, gaps_at = "Type")
```

Let's now show the __median__ methylation levels. Instead of plotting for each row an escapee CpGs, I aggregate CpGs by ensembl gene id __doing the median__ of the methylation levels. 

```{r, fig.width=12, fig.height=15}
MedianMeth <- aggregate(assays(bsseq_obj_subset)$Meth_WGBS, list(rowData(bsseq_obj_subset)$ensembl_gene_id), median, na.rm = T)
rownames(MedianMeth) <- MedianMeth$Group.1
MedianMeth <- MedianMeth[, -1]

rowDataSE <- rowData(bsseq_obj_subset)
rownames(rowDataSE) <- rowData(bsseq_obj_subset)$ensembl_gene_id
rowDataSE <- rowDataSE[rownames(MedianMeth), ]

SampleAnno <- colData(bsseq_obj_subset)

SE <- SummarizedExperiment(assays=MedianMeth, colData=SampleAnno, rowData = rowDataSE)

identical(colnames(assay(SE)), rownames(colData(SE))) #TRUE

# cellTypes_colors <- c(hiPSCs ="#dd1c77", iMeLCs ="#377eb8", hPGCLCs ="#4daf4a", hEGCLCs = "#ff7f00")
# colData(SE)$Type <- factor(colData(SE)$Type, levels = c("hiPSCs", "iMeLCs" , "hPGCLCs", "hEGCLCs"))
# 
# metadata(SE)$anno_colors <- list(Type = cellTypes_colors,
#                                                GenomicRegion = c ("3' UTR" = "#FFADAD", "5' UTR" = "#FFD6A5", "Distal Intergenic" = "#FDFFB6", "Downstream (<=300bp)" = "#FFFFFC", "Exon" = "#9BF6FF", "Intron" = "#A0C4FF", "Promoter (<=1kb)" = "#BDB2FF", "Promoter (1-2kb)" = "#FFC6FF", "Promoter (2-3kb)" = "#CAFFBF"))
```

```{r}
break_coord <- assay(SE) %>% quantile(seq(0,1, by=0.1), na.rm =TRUE)
ExpCols <- viridisLite::plasma(length(seq(head(break_coord,1), tail(break_coord,1), by=0.2)))
sechm(SE, features = rownames(SE)[1:3000], hmcols=ExpCols, breaks=seq(head(break_coord,1), tail(break_coord,1), by=0.2), show_rownames=FALSE, show_colnames=TRUE, do.scale=FALSE, column_title = "Median methylation levels (by gene) of CpGs in escapee regions", top_annotation = c("Type"), row_title_rot = 0, cluster_rows = TRUE, gaps_at = "Type")
```

Let's change the range of methylation (1st-10th deciles of methylation)

```{r}
break_coord <- assay(SE) %>% quantile(seq(0,1, by=0.1), na.rm =TRUE)
ExpCols <- viridisLite::plasma(length(seq(break_coord[2], tail(break_coord,1), by=0.05)))
sechm(SE, features = rownames(SE)[1:3000], hmcols=ExpCols, breaks=seq(break_coord[2], tail(break_coord,1), by=0.05), show_rownames=FALSE, show_colnames=TRUE, do.scale=FALSE, column_title = "Median methylation levels (by gene) of CpGs in escapee regions", top_annotation = c("Type"), row_title_rot = 0, cluster_rows = TRUE, gaps_at = "Type")
```

## 3.2 Only 5hmCs (bACEseq)

```{r, fig.width=12, fig.height=15}
AvgMeth <- aggregate(assays(bsseq_obj_subset)$Meth_hmC, list(rowData(bsseq_obj_subset)$ensembl_gene_id), mean, na.rm = T)
rownames(AvgMeth) <- AvgMeth$Group.1
AvgMeth <- AvgMeth[, -1]

rowDataSE <- rowData(bsseq_obj_subset)
rownames(rowDataSE) <- rowData(bsseq_obj_subset)$ensembl_gene_id
rowDataSE <- rowDataSE[rownames(AvgMeth), ]

SampleAnno <- colData(bsseq_obj_subset)

SE <- SummarizedExperiment(assays=AvgMeth, colData=SampleAnno, rowData = rowDataSE)

identical(colnames(assay(SE)), rownames(colData(SE))) #TRUE

# cellTypes_colors <- c(hiPSCs ="#dd1c77", iMeLCs ="#377eb8", hPGCLCs ="#4daf4a", hEGCLCs = "#ff7f00")
# colData(SE)$Type <- factor(colData(SE)$Type, levels = c("hiPSCs", "iMeLCs" , "hPGCLCs", "hEGCLCs"))
# 
# metadata(SE)$anno_colors <- list(Type = cellTypes_colors,
#                                                GenomicRegion = c ("3' UTR" = "#FFADAD", "5' UTR" = "#FFD6A5", "Distal Intergenic" = "#FDFFB6", "Downstream (<=300bp)" = "#FFFFFC", "Exon" = "#9BF6FF", "Intron" = "#A0C4FF", "Promoter (<=1kb)" = "#BDB2FF", "Promoter (1-2kb)" = "#FFC6FF", "Promoter (2-3kb)" = "#CAFFBF"))
```

Number of rows is `r length(AvgMeth)` therefore too many for an heatmap. I'll take only the first 3000.

```{r}
break_coord <- assay(SE) %>% quantile(seq(0,1, by=0.1), na.rm =TRUE)
ExpCols <- viridisLite::plasma(length(seq(head(break_coord,1), tail(break_coord,1), by=0.05)))
sechm(SE, features = rownames(SE)[1:3000], hmcols=ExpCols, breaks=seq(head(break_coord,1), tail(break_coord,1), by=0.05), show_rownames=FALSE, show_colnames=TRUE, do.scale=FALSE, column_title = "Average methylation levels (by gene) of CpGs in escapee regions", top_annotation = c("Type"), row_title_rot = 0, cluster_rows = TRUE, gaps_at = "Type")
```

Let's change the range of methylation

```{r}
break_coord <- assay(SE) %>% quantile(seq(0,1, by=0.1), na.rm =TRUE)
ExpCols <- viridisLite::plasma(length(seq(break_coord[2], tail(break_coord,1), by=0.05)))
sechm(SE, features = rownames(SE)[1:3000], hmcols=ExpCols, breaks=seq(break_coord[2], tail(break_coord,1), by=0.05), show_rownames=FALSE, show_colnames=TRUE, do.scale=FALSE, column_title = "Average methylation levels (by gene) of CpGs in escapee regions", top_annotation = c("Type"), row_title_rot = 0, cluster_rows = TRUE, gaps_at = "Type")
```

Let's now show the __median__ methylation levels. Instead of plotting for each row an escapee CpGs, I aggregate CpGs by ensembl gene id __doing the median__ of the methylation levels. 

```{r, fig.width=12, fig.height=15}
MedianMeth <- aggregate(assays(bsseq_obj_subset)$Meth_hmC, list(rowData(bsseq_obj_subset)$ensembl_gene_id), median, na.rm = T)
rownames(MedianMeth) <- MedianMeth$Group.1
MedianMeth <- MedianMeth[, -1]

rowDataSE <- rowData(bsseq_obj_subset)
rownames(rowDataSE) <- rowData(bsseq_obj_subset)$ensembl_gene_id
rowDataSE <- rowDataSE[rownames(MedianMeth), ]

SampleAnno <- colData(bsseq_obj_subset)

SE <- SummarizedExperiment(assays=MedianMeth, colData=SampleAnno, rowData = rowDataSE)

identical(colnames(assay(SE)), rownames(colData(SE))) #TRUE

# cellTypes_colors <- c(hiPSCs ="#dd1c77", iMeLCs ="#377eb8", hPGCLCs ="#4daf4a", hEGCLCs = "#ff7f00")
# colData(SE)$Type <- factor(colData(SE)$Type, levels = c("hiPSCs", "iMeLCs" , "hPGCLCs", "hEGCLCs"))
# 
# metadata(SE)$anno_colors <- list(Type = cellTypes_colors,
#                                                GenomicRegion = c ("3' UTR" = "#FFADAD", "5' UTR" = "#FFD6A5", "Distal Intergenic" = "#FDFFB6", "Downstream (<=300bp)" = "#FFFFFC", "Exon" = "#9BF6FF", "Intron" = "#A0C4FF", "Promoter (<=1kb)" = "#BDB2FF", "Promoter (1-2kb)" = "#FFC6FF", "Promoter (2-3kb)" = "#CAFFBF"))
```

```{r}
break_coord <- assay(SE) %>% quantile(seq(0,1, by=0.1), na.rm =TRUE)
ExpCols <- viridisLite::plasma(length(seq(head(break_coord,1), tail(break_coord,1), by=0.05)))
sechm(SE, features = rownames(SE)[1:3000], hmcols=ExpCols, breaks=seq(head(break_coord,1), tail(break_coord,1), by=0.05), show_rownames=FALSE, show_colnames=TRUE, do.scale=FALSE, column_title = "Median methylation levels (by gene) of CpGs in escapee regions", top_annotation = c("Type"), row_title_rot = 0, cluster_rows = TRUE, gaps_at = "Type")
```

Let's change the range of methylation (1st-10th deciles of methylation)

```{r}
break_coord <- assay(SE) %>% quantile(seq(0,1, by=0.1), na.rm =TRUE)
ExpCols <- viridisLite::plasma(length(seq(break_coord[2], tail(break_coord,1), by=0.05)))
sechm(SE, features = rownames(SE)[1:3000], hmcols=ExpCols, breaks=seq(break_coord[2], tail(break_coord,1), by=0.05), show_rownames=FALSE, show_colnames=TRUE, do.scale=FALSE, column_title = "Median methylation levels (by gene) of CpGs in escapee regions", top_annotation = c("Type"), row_title_rot = 0, cluster_rows = TRUE, gaps_at = "Type")
```

## 3.3 Only 5mCs (WGBS - bACEseq)

```{r, fig.width=12, fig.height=15}
AvgMeth <- aggregate(assays(bsseq_obj_subset)$Meth_mC, list(rowData(bsseq_obj_subset)$ensembl_gene_id), mean, na.rm = T)
rownames(AvgMeth) <- AvgMeth$Group.1
AvgMeth <- AvgMeth[, -1]

rowDataSE <- rowData(bsseq_obj_subset)
rownames(rowDataSE) <- rowData(bsseq_obj_subset)$ensembl_gene_id
rowDataSE <- rowDataSE[rownames(AvgMeth), ]

SampleAnno <- colData(bsseq_obj_subset)

SE <- SummarizedExperiment(assays=AvgMeth, colData=SampleAnno, rowData = rowDataSE)

identical(colnames(assay(SE)), rownames(colData(SE))) #TRUE

# cellTypes_colors <- c(hiPSCs ="#dd1c77", iMeLCs ="#377eb8", hPGCLCs ="#4daf4a", hEGCLCs = "#ff7f00")
# colData(SE)$Type <- factor(colData(SE)$Type, levels = c("hiPSCs", "iMeLCs" , "hPGCLCs", "hEGCLCs"))
# 
# metadata(SE)$anno_colors <- list(Type = cellTypes_colors,
#                                                GenomicRegion = c ("3' UTR" = "#FFADAD", "5' UTR" = "#FFD6A5", "Distal Intergenic" = "#FDFFB6", "Downstream (<=300bp)" = "#FFFFFC", "Exon" = "#9BF6FF", "Intron" = "#A0C4FF", "Promoter (<=1kb)" = "#BDB2FF", "Promoter (1-2kb)" = "#FFC6FF", "Promoter (2-3kb)" = "#CAFFBF"))
```

Number of rows is `r length(AvgMeth)` therefore too many for an heatmap. I'll take only the first 3000.

```{r}
break_coord <- assay(SE) %>% quantile(seq(0,1, by=0.1), na.rm =TRUE)
ExpCols <- viridisLite::plasma(length(seq(head(break_coord,1), tail(break_coord,1), by=0.2)))
sechm(SE, features = rownames(SE)[1:3000], hmcols=ExpCols, breaks=seq(head(break_coord,1), tail(break_coord,1), by=0.2), show_rownames=FALSE, show_colnames=TRUE, do.scale=FALSE, column_title = "Average methylation levels (by gene) of CpGs in escapee regions", top_annotation = c("Type"), row_title_rot = 0, cluster_rows = TRUE, gaps_at = "Type")
```

Let's change the range of methylation

```{r}
break_coord <- assay(SE) %>% quantile(seq(0,1, by=0.1), na.rm =TRUE)
ExpCols <- viridisLite::plasma(length(seq(break_coord[2], tail(break_coord,1), by=0.05)))
sechm(SE, features = rownames(SE)[1:3000], hmcols=ExpCols, breaks=seq(break_coord[2], tail(break_coord,1), by=0.05), show_rownames=FALSE, show_colnames=TRUE, do.scale=FALSE, column_title = "Average methylation levels (by gene) of CpGs in escapee regions", top_annotation = c("Type"), row_title_rot = 0, cluster_rows = TRUE, gaps_at = "Type")
```

Let's now show the __median__ methylation levels. Instead of plotting for each row an escapee CpGs, I aggregate CpGs by ensembl gene id __doing the median__ of the methylation levels. 

```{r, fig.width=12, fig.height=15}
MedianMeth <- aggregate(assays(bsseq_obj_subset)$Meth_mC, list(rowData(bsseq_obj_subset)$ensembl_gene_id), median, na.rm = T)
rownames(MedianMeth) <- MedianMeth$Group.1
MedianMeth <- MedianMeth[, -1]

rowDataSE <- rowData(bsseq_obj_subset)
rownames(rowDataSE) <- rowData(bsseq_obj_subset)$ensembl_gene_id
rowDataSE <- rowDataSE[rownames(MedianMeth), ]

SampleAnno <- colData(bsseq_obj_subset)

SE <- SummarizedExperiment(assays=MedianMeth, colData=SampleAnno, rowData = rowDataSE)

identical(colnames(assay(SE)), rownames(colData(SE))) #TRUE

# cellTypes_colors <- c(hiPSCs ="#dd1c77", iMeLCs ="#377eb8", hPGCLCs ="#4daf4a", hEGCLCs = "#ff7f00")
# colData(SE)$Type <- factor(colData(SE)$Type, levels = c("hiPSCs", "iMeLCs" , "hPGCLCs", "hEGCLCs"))
# 
# metadata(SE)$anno_colors <- list(Type = cellTypes_colors,
#                                                GenomicRegion = c ("3' UTR" = "#FFADAD", "5' UTR" = "#FFD6A5", "Distal Intergenic" = "#FDFFB6", "Downstream (<=300bp)" = "#FFFFFC", "Exon" = "#9BF6FF", "Intron" = "#A0C4FF", "Promoter (<=1kb)" = "#BDB2FF", "Promoter (1-2kb)" = "#FFC6FF", "Promoter (2-3kb)" = "#CAFFBF"))
```

```{r}
break_coord <- assay(SE) %>% quantile(seq(0,1, by=0.1), na.rm =TRUE)
ExpCols <- viridisLite::plasma(length(seq(head(break_coord,1), tail(break_coord,1), by=0.2)))
sechm(SE, features = rownames(SE)[1:3000], hmcols=ExpCols, breaks=seq(head(break_coord,1), tail(break_coord,1), by=0.2), show_rownames=FALSE, show_colnames=TRUE, do.scale=FALSE, column_title = "Median methylation levels (by gene) of CpGs in escapee regions", top_annotation = c("Type"), row_title_rot = 0, cluster_rows = TRUE, gaps_at = "Type")
```

Let's change the range of methylation (1st-10th deciles of methylation)

```{r}
break_coord <- assay(SE) %>% quantile(seq(0,1, by=0.1), na.rm =TRUE)
ExpCols <- viridisLite::plasma(length(seq(break_coord[2], tail(break_coord,1), by=0.05)))
sechm(SE, features = rownames(SE)[1:3000], hmcols=ExpCols, breaks=seq(break_coord[2], tail(break_coord,1), by=0.05), show_rownames=FALSE, show_colnames=TRUE, do.scale=FALSE, column_title = "Median methylation levels (by gene) of CpGs in escapee regions", top_annotation = c("Type"), row_title_rot = 0, cluster_rows = TRUE, gaps_at = "Type")
```

# 4. Session Info

```{r}
date()
sessionInfo()
```

