---
title: 'Comparison escapees vs non-escapees - Hsu et al.'
author: 'Gaja Matassa'
date: 'Date: `r format(Sys.Date(), "%B %d, %Y")`'
output:
    html_document:
        toc: true
        toc_float: true
        theme: 'united'
        highlight: 'kate'
        code_folding: hide
params: 
    InputFolder: '~/DataDir/7.EscapeesExploration/Output/EscapeeBsseq_Hsu/'
    num_permutations: 500
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, warning = FALSE)
```

```{r}
library(plyr)
library(dplyr)
library(bsseq)
library(dmrseq)
library(ggplot2)
source("../Methylation_helper.R")
```

Set random seed and number of permutations
```{r}
set.seed(123)
num_permutations <- params$num_permutations
```

# 1. Data loading and fast exploration by PCA

Loading bsseq objects

```{r}
bsseq_obj_escapees <- readRDS(paste0(params$InputFolder, "bsseq_obj_escapees_Hsu.rds"))
bsseq_obj_NonEscapees <- readRDS(paste0(params$InputFolder, "bsseq_obj_NonEscapees_Hsu.rds"))
```

* `r dim(bsseq_obj_escapees)[1]` CpG loci belong to escapees regions (escapee CpGs).
* `r dim(bsseq_obj_NonEscapees)[1]` CpG loci do not belong to escapees regions (non-escapee CpGs).

**Colors for figures**

```{r}
cellTypes_colors <- c(hESC ="#dd1c77", D4hPGCLC ="#4daf4a")
```

## PCA only on escapee CpGs - For now focus on WGBS

```{r, collapse=TRUE, fig.width=8}
pca_res <- do_PCA(assays(bsseq_obj_escapees)$Meth_WGBS)
sample_anno <- bsseq::pData(bsseq_obj_escapees)
plot_PCA(pca_res = pca_res, anno = sample_anno, col_anno = "Type", shape_anno = "Line", custom_colors = cellTypes_colors, point_size = 5)
```

## PCA only on non-escapee CpGs

```{r, collapse=TRUE, fig.width=8}
pca_res <- do_PCA(assays(bsseq_obj_NonEscapees)$Meth_WGBS)
sample_anno <- bsseq::pData(bsseq_obj_NonEscapees)
plot_PCA(pca_res = pca_res, anno = sample_anno, col_anno = "Type", shape_anno = "Line", custom_colors = cellTypes_colors, point_size = 5)
```

# 2.Generation of `r num_permutations` random sets of `r dim(bsseq_obj_escapees)[1]` non-escapee CpGs and distributions plots

Let's generate random indices to randomly select `r dim(bsseq_obj_escapees)[1]` CpG loci among the non-escapee ones.

```{r}
random_sets <- list()

for (i in 1:num_permutations) {
  random_indices <- sample(1:dim(bsseq_obj_NonEscapees)[1], dim(bsseq_obj_escapees)[1], replace = FALSE)
  bsseq_obj_randomNonEscapees <- bsseq_obj_NonEscapees[random_indices, ]
  
  random_sets[[i]] <- bsseq_obj_randomNonEscapees
}
```

**Head of BSSEQ with escapee CpGs**

```{r, collapse=FALSE}
assays(bsseq_obj_escapees)$Meth_WGBS %>% head()
```

**Escapee CpG methylation distribution plot**

```{r}
hist(assays(bsseq_obj_escapees)$Meth_WGB, breaks = 100, main = "Distribution of Methylation (Meth_WGBS)",
     xlab = "Methylation Level", col = "skyblue", border = "white")
```

```{r}
assays(bsseq_obj_escapees)$M <- assays(bsseq_obj_escapees)$M_WGBS
assays(bsseq_obj_escapees)$Cov <- assays(bsseq_obj_escapees)$Cov_WGBS
plotEmpiricalDistribution(bsseq_obj_escapees, 
                          bySample = TRUE,
                          testCovariate = "Type",
                          adj = 3) + guides(linetype="none") + 
  labs(title = "Escapee CpG methylation distribution by cell type") + ggplot2::scale_color_manual(values = cellTypes_colors)
```

```{r}
plotEmpiricalDistribution(bsseq_obj_escapees, 
                          bySample = FALSE,
                          testCovariate = "Line",
                          adj = 3) + guides(linetype="none") + 
  labs(title = "Escapee CpG methylation distribution collapsing cell types")
```

**Non-Escapee CpG methylation distribution plot**

```{r}
assays(bsseq_obj_NonEscapees)$M <- assays(bsseq_obj_NonEscapees)$M_WGBS
assays(bsseq_obj_NonEscapees)$Cov <- assays(bsseq_obj_NonEscapees)$Cov_WGBS
plotEmpiricalDistribution(bsseq_obj_NonEscapees, 
                          bySample = TRUE,
                          testCovariate = "Type",
                          adj = 3) + guides(linetype="none") + 
  labs(title = "Non-escapee CpG methylation distribution by cell type") + ggplot2::scale_color_manual(values = cellTypes_colors)
```

```{r}
plotEmpiricalDistribution(bsseq_obj_NonEscapees, 
                          bySample = FALSE,
                          testCovariate = "Line",
                          adj = 3) + guides(linetype="none") + 
  labs(title = "Non-escapee CpG methylation distribution collapsing cell types")
```

**Head of first BSSEQ with randomly selected non-escapee CpGs**

```{r, collapse=FALSE}
assays(random_sets[[1]])$Meth_WGBS %>% head()
```

**First set of randomly selected non-escapee CpG methylation distribution plot**

```{r}
assays(random_sets[[1]])$M <- assays(random_sets[[1]])$M_WGBS
assays(random_sets[[1]])$Cov <- assays(random_sets[[1]])$Cov_WGBS
plotEmpiricalDistribution(random_sets[[1]], 
                          bySample = TRUE,
                          testCovariate = "Type",
                          adj = 3) + guides(linetype="none") + 
  labs(title = "Non-escapee CpG methylation distribution by cell type (1)") + ggplot2::scale_color_manual(values = cellTypes_colors)
```

**Head of second BSSEQ with randomly selected non-escapee CpGs**

```{r, collapse=FALSE}
assays(random_sets[[2]])$Meth_WGBS %>% head()
```

**Second set of randomly selected non-escapee CpG methylation distribution plot**

```{r}
assays(random_sets[[2]])$M <- assays(random_sets[[2]])$M_WGBS
assays(random_sets[[2]])$Cov <- assays(random_sets[[2]])$Cov_WGBS
plotEmpiricalDistribution(random_sets[[2]], 
                          bySample = TRUE,
                          testCovariate = "Type",
                          adj = 3) + guides(linetype="none") + 
  labs(title = "Non-escapee CpG methylation distribution by cell type (2)") + ggplot2::scale_color_manual(values = cellTypes_colors)
```

# 3. Kolmogorov-Smirnov test

Example of KS test result comparing distributions in case of UCLA1_hESC_rep3 sample: escapee CpG methylation levels vs first randomly selected non-escapee set.

```{r, collapse=FALSE}
ks_test_result <- ks.test(assays(bsseq_obj_escapees)$Meth_WGBS[, "UCLA1_hESC_rep3"], assays(random_sets[[1]])$Meth_WGBS[, "UCLA1_hESC_rep3"])
print(ks_test_result)
```

Example of KS test result comparing distributions in case of UCLA1_hESC_rep3 sample: first randomly selected non-escapee set vs second randomly selected non-escapee set.

```{r, collapse=FALSE}
ks_test_result <- ks.test(assays(random_sets[[1]])$Meth_WGBS[, "UCLA1_hESC_rep3"], assays(random_sets[[2]])$Meth_WGBS[, "UCLA1_hESC_rep3"])
print(ks_test_result)
```

Example of KS test result comparing distributions: escapee CpG methylation levels (UCLA2_hESC_rep2 sample) vs first randomly selected non-escapee set (UCLA1_D4hPGCLC_rep3 sample).

```{r, collapse=FALSE}
ks_test_result <- ks.test(assays(bsseq_obj_escapees)$Meth_WGBS[, "UCLA2_hESC_rep2"],  assays(random_sets[[1]])$Meth_WGBS[, "UCLA1_D4hPGCLC_rep3"])
print(ks_test_result)
```

Example of KS test result comparing distributions: escapee CpG methylation levels vs first randomly selected non-escapee set **considering the mean levels across cell types (rowwise)**.

```{r, collapse=FALSE}
ks_test_result <- ks.test(assays(bsseq_obj_escapees)$Meth_WGBS %>% rowMeans(na.rm = TRUE), assays(random_sets[[1]])$Meth_WGBS %>% rowMeans(na.rm = TRUE))
print(ks_test_result)
```

Example of KS test result comparing distributions: first randomly selected non-escapee set vs second randomly selected non-escapee set **considering the mean levels across cell types (rowwise)**.

```{r, collapse=FALSE}
ks_test_result <- ks.test(assays(random_sets[[1]])$Meth_WGBS %>% rowMeans(na.rm = TRUE), assays(random_sets[[2]])$Meth_WGBS %>% rowMeans(na.rm = TRUE))
print(ks_test_result)
```

*Let's calculate the mean methylation levels across cell types in all sets of randomly selected non-escapee CpGs.*

```{r}
random_MeanMethLevels <- lapply(random_sets, function(bsseq){
  assays(bsseq)$Meth_WGBS %>% rowMeans(na.rm = TRUE)})
```

**Let's do KS test comparing escapee CpG methylation levels vs all `r num_permutations` randomly selected non-escapee sets.**
```{r, collapse=FALSE}
ks_test_results <- lapply(random_MeanMethLevels, function(random_MeanMeth) {
  ks.test(assays(bsseq_obj_escapees)$Meth_WGBS %>% rowMeans(na.rm = TRUE), random_MeanMeth)
})
pvalues_kstest <- sapply(ks_test_results, function(result) {
  result$p.value
})
adjusted_pvalues_kstest <- p.adjust(pvalues_kstest, method = "fdr")
print(adjusted_pvalues_kstest)
```

How many adjusted p-values are < 0.05?

```{r, collapse=FALSE}
table(adjusted_pvalues_kstest < 0.05)
```

**Let's do KS test comparing first randomly selected non-escapee set vs all `r num_permutations` randomly selected non-escapee sets.**

```{r, collapse=FALSE}
ks_test_results <- lapply(random_MeanMethLevels, function(random_MeanMeth) {
  ks.test(random_MeanMethLevels[[1]], random_MeanMeth)
})
pvalues_kstest <- sapply(ks_test_results, function(result) {
  result$p.value
})
adjusted_pvalues_kstest <- p.adjust(pvalues_kstest, method = "fdr")
print(adjusted_pvalues_kstest)
```

How many adjusted p-values are < 0.05?

```{r, collapse=FALSE}
table(adjusted_pvalues_kstest < 0.05)
```

# 4.Escapee and non escapee CpG methylation distribution plot 

## 4.1 WGBS Meth

```{r}
MeanMethLevels_df <- c(unlist(random_MeanMethLevels), assays(bsseq_obj_escapees)$Meth_WGBS %>% rowMeans(na.rm = TRUE) %>% as.vector()) %>% as.data.frame()
colnames(MeanMethLevels_df) <- "MeanMeth"
MeanMethLevels_df$Distribution <- c(paste0("Random_", rep(1:num_permutations, each = dim(bsseq_obj_escapees)[1])), rep("Escapees_CpGs", each = dim(bsseq_obj_escapees)[1]))
MeanMethLevels_df$Category <- "Non Escapees"
MeanMethLevels_df$Category[(nrow(MeanMethLevels_df)+1-nrow(bsseq_obj_escapees)):nrow(MeanMethLevels_df)] <- "Escapees"
```

### Mean methylation levels collapsing cell types

```{r, fig.width=10}
ggplot(MeanMethLevels_df, aes(x = MeanMeth, color = Category, group = Distribution)) +
  geom_density(alpha = 0.5)  +
  labs(
    title = "Distribution CpG mean methylation levels in escapees and non-escapees regions", 
    x = "Mean Methylation Levels across cell types", 
    y = "Density", 
    color = "Category"
  ) +
  scale_color_manual(values = c("Escapees" = "#FF7D00", "Non Escapees" = "#15616D")) + 
  theme_minimal()
```

### Mean methylation levels separating cell types

```{r}
# Define function to compute rowMeans for specific sample types
get_rowMeans_by_type <- function(bsseq, type_label) {
  sample_indices <- which(colData(bsseq)$Type == type_label)
  rowMeans(assays(bsseq)$Meth_WGBS[, sample_indices, drop = FALSE], na.rm = TRUE)
}

# Apply the function to each random set for both hESC and D4hPGCLC
random_MeanMethLevels_hESC <- lapply(random_sets, get_rowMeans_by_type, type_label = "hESC")
random_MeanMethLevels_D4hPGCLC <- lapply(random_sets, get_rowMeans_by_type, type_label = "D4hPGCLC")

# Do the same for escapees
escapees_MeanMeth_hESC <- get_rowMeans_by_type(bsseq_obj_escapees, "hESC")
escapees_MeanMeth_D4hPGCLC <- get_rowMeans_by_type(bsseq_obj_escapees, "D4hPGCLC")

# Combine all into a single data frame
MeanMethLevels_df <- data.frame(
  MeanMeth = c(
    unlist(random_MeanMethLevels_hESC),
    unlist(random_MeanMethLevels_D4hPGCLC),
    escapees_MeanMeth_hESC,
    escapees_MeanMeth_D4hPGCLC
  ),
  Distribution = c(
    paste0("Random_hESC_", rep(1:num_permutations, each = dim(bsseq_obj_escapees)[1])),
    paste0("Random_D4hPGCLC_", rep(1:num_permutations, each = dim(bsseq_obj_escapees)[1])),
    rep("Escapees_hESC", each = dim(bsseq_obj_escapees)[1]),
    rep("Escapees_D4hPGCLC", each = dim(bsseq_obj_escapees)[1])
  ),
  Category = c(
    rep("Non Escapees", 2 * num_permutations * dim(bsseq_obj_escapees)[1]),
    rep("Escapees", 2 * dim(bsseq_obj_escapees)[1])
  ),
  Type = c(
    rep("hESC", num_permutations * dim(bsseq_obj_escapees)[1]),
    rep("D4hPGCLC", num_permutations * dim(bsseq_obj_escapees)[1]),
    rep("hESC", dim(bsseq_obj_escapees)[1]),
    rep("D4hPGCLC", dim(bsseq_obj_escapees)[1])
  )
)

ggplot(MeanMethLevels_df, aes(x = MeanMeth, color = Category, group = Distribution)) +
  geom_density(alpha = 0.5)  +
  labs(
    title = "CpG Mean Methylation Levels by Category and Type", 
    x = "Mean Methylation Levels", 
    y = "Density", 
    color = "Category"
  ) +
  facet_wrap(~ Type, scales = "free_y") +
  scale_color_manual(values = c("Escapees" = "#FF7D00", "Non Escapees" = "#15616D")) + 
  theme_minimal()
```

```{r}
# Create a new combined variable in the dataframe
MeanMethLevels_df$GroupLabel <- paste(MeanMethLevels_df$Category, MeanMethLevels_df$Type, sep = "_")

# Now plot using this combined label
ggplot(MeanMethLevels_df, aes(x = MeanMeth, color = GroupLabel, group = GroupLabel)) +
  geom_density(alpha = 0.6, adjust = 1.5) +  # smoother curves using `adjust`
  labs(
    title = "CpG Mean Methylation Levels by Category and Cell Type",
    x = "Mean Methylation Levels",
    y = "Density",
    color = "Group"
  ) +
  scale_color_manual(
    values = c(
      "Escapees_hESC" = "#FF7D00",
      "Escapees_D4hPGCLC" = "#FFB347",
      "Non Escapees_hESC" = "#15616D",
      "Non Escapees_D4hPGCLC" = "#1E91A1"
    )
  ) +
  coord_cartesian(ylim = c(0, 6)) +  # cut off at y = 6
  theme_minimal()
```

## 4.2 bACEseq Meth

```{r}
random_MeanHydroxyMethLevels <- lapply(random_sets, function(bsseq){
  assays(bsseq)$Meth_hmC %>% rowMeans(na.rm = TRUE)})
MeanHydroxyMethLevels_df <- c(unlist(random_MeanHydroxyMethLevels), assays(bsseq_obj_escapees)$Meth_hmC %>% rowMeans(na.rm = TRUE) %>% as.vector()) %>% as.data.frame()
colnames(MeanHydroxyMethLevels_df) <- "MeanHydroxyMeth"
MeanHydroxyMethLevels_df$Distribution <- c(paste0("Random_", rep(1:num_permutations, each = dim(bsseq_obj_escapees)[1])), rep("Escapees_CpGs", each = dim(bsseq_obj_escapees)[1]))
MeanHydroxyMethLevels_df$Category <- "Non Escapees"
MeanHydroxyMethLevels_df$Category[(nrow(MeanHydroxyMethLevels_df)+1-nrow(bsseq_obj_escapees)):nrow(MeanHydroxyMethLevels_df)] <- "Escapees"
```

```{r, fig.width=10}
ggplot(MeanHydroxyMethLevels_df, aes(x = MeanHydroxyMeth, color = Category, group = Distribution)) +
  geom_density(alpha = 0.5)  +
  labs(
    title = "Distribution CpG mean hydroxymethylation levels in escapees and non-escapees regions", 
    x = "Mean Hydroxymethylation Levels across cell types", 
    y = "Density", 
    color = "Category"
  ) +
  scale_color_manual(values = c("Escapees" = "#FF7D00", "Non Escapees" = "#15616D"))  + 
  coord_cartesian(xlim = c(0, 0.1)) +
  theme_minimal()
```

```{r}
# Define function to compute rowMeans for specific sample types
get_rowMeans_by_type_hydroxy <- function(bsseq, type_label) {
  sample_indices <- which(colData(bsseq)$Type == type_label)
  rowMeans(assays(bsseq)$Meth_hmC[, sample_indices, drop = FALSE], na.rm = TRUE)
}

# Apply the function to each random set for both hESC and D4hPGCLC
random_MeanHydroxyMethLevels_hESC <- lapply(random_sets, get_rowMeans_by_type_hydroxy, type_label = "hESC")
random_MeanHydroxyMethLevels_D4hPGCLC <- lapply(random_sets, get_rowMeans_by_type_hydroxy, type_label = "D4hPGCLC")

# Do the same for escapees
escapees_MeanHydroxyMeth_hESC <- get_rowMeans_by_type_hydroxy(bsseq_obj_escapees, "hESC")
escapees_MeanHydroxyMeth_D4hPGCLC <- get_rowMeans_by_type_hydroxy(bsseq_obj_escapees, "D4hPGCLC")

# Combine all into a single data frame
MeanHydroxyMethLevels_df <- data.frame(
  MeanHydroxyMeth = c(
    unlist(random_MeanHydroxyMethLevels_hESC),
    unlist(random_MeanHydroxyMethLevels_D4hPGCLC),
    escapees_MeanHydroxyMeth_hESC,
    escapees_MeanHydroxyMeth_D4hPGCLC
  ),
  Distribution = c(
    paste0("Random_hESC_", rep(1:num_permutations, each = dim(bsseq_obj_escapees)[1])),
    paste0("Random_D4hPGCLC_", rep(1:num_permutations, each = dim(bsseq_obj_escapees)[1])),
    rep("Escapees_hESC", each = dim(bsseq_obj_escapees)[1]),
    rep("Escapees_D4hPGCLC", each = dim(bsseq_obj_escapees)[1])
  ),
  Category = c(
    rep("Non Escapees", 2 * num_permutations * dim(bsseq_obj_escapees)[1]),
    rep("Escapees", 2 * dim(bsseq_obj_escapees)[1])
  ),
  Type = c(
    rep("hESC", num_permutations * dim(bsseq_obj_escapees)[1]),
    rep("D4hPGCLC", num_permutations * dim(bsseq_obj_escapees)[1]),
    rep("hESC", dim(bsseq_obj_escapees)[1]),
    rep("D4hPGCLC", dim(bsseq_obj_escapees)[1])
  )
)

ggplot(MeanHydroxyMethLevels_df, aes(x = MeanHydroxyMeth, color = Category, group = Distribution)) +
  geom_density(alpha = 0.5)  +
  labs(
    title = "CpG Mean Hydroxy Methylation Levels by Category and Type", 
    x = "Mean Methylation Levels", 
    y = "Density", 
    color = "Category"
  ) +
  facet_wrap(~ Type, scales = "free_y") +
  scale_color_manual(values = c("Escapees" = "#FF7D00", "Non Escapees" = "#15616D")) + 
  theme_minimal()
```

```{r}
# Create a new combined variable in the dataframe
MeanHydroxyMethLevels_df$GroupLabel <- paste(MeanHydroxyMethLevels_df$Category, MeanHydroxyMethLevels_df$Type, sep = "_")

# Now plot using this combined label
ggplot(MeanHydroxyMethLevels_df, aes(x = MeanHydroxyMeth, color = GroupLabel, group = GroupLabel)) +
  geom_density(alpha = 0.6, adjust = 1.5) +  # smoother curves using `adjust`
  labs(
    title = "Distribution CpG mean hydroxymethylation levels in escapees and non-escapees regions",
    x = "Mean Hydroxymethylation Levels",
    y = "Density",
    color = "Group"
  ) +
  scale_color_manual(
    values = c(
      "Escapees_hESC" = "#FF7D00",
      "Escapees_D4hPGCLC" = "#FFB347",
      "Non Escapees_hESC" = "#15616D",
      "Non Escapees_D4hPGCLC" = "#1E91A1"
    )
  ) +
  coord_cartesian(xlim = c(0, 0.25), ylim = c(0, 60)) +  # cut off at y = 6
  theme_minimal()
```

# 5.Session Info

```{r}
date()
sessionInfo()
```

