---
title: "DMR Annotation"
author: "Sarah Stucchi"
date: 'Date: `r format(Sys.Date(), "%B %d, %Y")`'
output:
    html_document:
        toc: true
        toc_float: true
        theme: 'united'
        highlight: 'kate'
        code_folding: hide
params: 
    InputFolder: '/group/testa/Project/PGCLC/Rebuttal_paperEG2025/DM/'
    OutputFolder: '/group/testa/Project/PGCLC/Rebuttal_paperEG2025/4_DMRAnnotated/Rev'
    TwistAnnotated: '/group/testa/Project/PGCLC/NewData_fromLaSitua/3.TwistBedAnn/TwistAnnotated.rds'
---

# 1. Environment setting

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE)
#knitr::opts_chunk$set(echo = TRUE, warning=FALSE, dev = 'cairo_pdf')

```

```{r, collapse=TRUE}
library(DSS)
library(plotly)
library(ggplot2)
library(scales)
library(gridExtra)
library(ChIPseeker)
library(org.Hs.eg.db)
library(ReactomePA)
library(clusterProfiler)
library(dmrseq)
library(annotatr)
library(DT)
library(data.table)

GTFFile <- "/group/testa/Project/PGCLC/NewData_fromLaSitua/3.TwistBedAnn/Input/gencode.v35.annotation.gtf.gz"
txdb_v35 <- GenomicFeatures::makeTxDbFromGFF(GTFFile, format="gtf")
```

```{r, collapse=TRUE}
InputFolder <- params$InputFolder
OutputFolder <- params$OutputFolder
```

# 2. Loading data

## 2.1 BSSeq objects

Loading of bsseq object, with CpGs shared by 75% of samples (after sample selection).

```{r, collapse=TRUE}
bsseq_obj <- readRDS("/group/testa/Project/PGCLC/emseq_rebuttal/2.PreliminaryAnalysis/bsseq_obj_sharedby75ofall.rds") #getting bsseq object where sample selection and CpG filtering were already performed   
```

## 2.2 DMRs and DMLs

**The DMRs here selected are called like this: `delta=0.25` and `p.threshold=0.0001`** 

```{r}
DMRs <- readRDS(paste0(InputFolder, '/', 'DMRs.rds'))
#DMLs <- readRDS(paste0(InputFolder, '/', 'DMLs.rds'))
```

## 2.3 Universe of genes

Importing Twist regions that were annotated
```{r}
TwistAnnotated <- readRDS(params$TwistAnnotated)
TwistAnnotated %>% head()
```

```{r}
Genes_df <- TwistAnnotated[, c("ensembl_gene_id_version", "ensembl_gene_id", "hgnc_symbol", "gene_biotype")] %>% dplyr::distinct()%>%filter(!is.na(hgnc_symbol))%>%filter(!hgnc_symbol%in%"")
rownames(Genes_df) <- NULL
```

These are the genes with the same symbol but different ensembl id

```{r}
Genes_df[Genes_df$hgnc_symbol%in%Genes_df[duplicated(Genes_df$hgnc_symbol), "hgnc_symbol"],]
```

```{r}
Genes_df$entrez_gene_id <- as.vector(mapIds(org.Hs.eg.db, Genes_df$hgnc_symbol, "ENTREZID","SYMBOL"))
```

```{r}
nrow(Genes_df[is.na(Genes_df$entrez_gene_id),]) #number of genes without an entrez id
```

```{r}
GeneUniverse_symbol <- unique(Genes_df$hgnc_symbol)
GeneUniverse_entrezid <- na.omit(unique(Genes_df$entrez_gene_id))
```

# 3. Transform DataFrame with DMRs info to GRanges object 

Remember that position form bedGraph files are 0-based while R works with 1-based position and therefore also the packages we will go to use, this is the reason why I set `starts.in.df.are.0based = TRUE` in creating the GRanges objects, it converts 0-based to 1-based coordinates.

```{r}
DMRs_GRanges <- DMRs
#DMLs_GRanges <- DMLs

for(i in 1:length(DMRs)){
  DMRs_GRanges[[i]] <- makeGRangesFromDataFrame(DMRs_GRanges[[i]], keep.extra.columns = TRUE, 
                                                start.field = "start", end.field = "end", starts.in.df.are.0based = TRUE)
#DMLs_GRanges[[i]] <- makeGRangesFromDataFrame(DMLs_GRanges[[i]], keep.extra.columns = TRUE, 
  #start.field = "start", end.field = "end", starts.in.df.are.0based = TRUE)
  } 
```


* For hiPSCs vs hEGCLC_p10 we have `r nrow(DMRs$hEGCLCs_p10vshiPSCs)` DMRs.
* For iMeLCs vs hEGCLCs_p10 we have `r nrow(DMRs$hEGCLCs_p10vsiMeLCs)` DMRs.
* For hPGCLCs vs hEGCLC_p10 we have `r nrow(DMRs$hEGCLCs_p10vshPGCLCs)` DMRs.
* For hEGCLCs_p5 vs hEGCLCs_p10 we have `r nrow(DMRs$hEGCLCs_p10vshEGCLCs_p5)` DMRs.

# 4. DMRs annotation {.tabset}



## 4.1 hiPSCs vs hEGCLCs_p10

### 4.1.1 Genetic annotation and visualization with __`ChIPseeker`__

DMRs are now annotated to their associated genes and genomic regions using the `annotatePeaks` function of the `ChIPseeker `R package, using a TxDb object. The [`GenomicFeatures`](https://bioconductor.org/packages/devel/bioc/vignettes/GenomicFeatures/inst/doc/GenomicFeatures.html) package uses `TxDb` objects to store transcript metadata. This class maps the 5’ and 3’ untranslated regions (UTRs), protein coding sequences (CDSs) and exons for a set of mRNA transcripts to their associated genome. 
All `TxDb` objects are backed by a SQLite database that manages genomic locations and the relationships between pre-processed mRNA transcripts, exons, protein coding sequences, and their related gene identifiers.

**TxDb created by us**

```{r, collapse=TRUE}
peakAnnoList <- lapply(list("Higher methylation in hiPSCs" = DMRs_GRanges$hEGCLCs_p10vshiPSCs[DMRs_GRanges$hEGCLCs_p10vshiPSCs$diff.Methy<0,], "Higher methylation in hEGCLCs_p10" = DMRs_GRanges$hEGCLCs_p10vshiPSCs[DMRs_GRanges$hEGCLCs_p10vshiPSCs$diff.Methy>0,]), annotatePeak, TxDb=txdb_v35, tssRegion=c(-3000, 3000), verbose=FALSE)
```

```{r, collapse=TRUE}
ChIPseeker::plotAnnoBar(peakAnnoList)
```

```{r, collapse=TRUE}
ChIPseeker::plotDistToTSS(peakAnnoList)
```

Given a list of gene set, `compareCluster` function will compute profiles of each gene cluster. 

```{r, collapse=TRUE}
genes = lapply(peakAnnoList, function(i) unique(as.data.frame(i)$geneId))
names(genes) = sub("_", "\n", names(genes))
```

```{r}
any(duplicated(genes$`Higher methylation in hiPSCs`))
```

```{r}
any(duplicated(genes$`Higher methylation in hEGCLCs_p10`))
```

There are no duplicated ensembl ids

```{r}
table(genes$`Higher methylation in hEGCLCs_p10` %in% Genes_df$ensembl_gene_id_version)
```

```{r}
genes$`Higher methylation in hEGCLCs_p10` [!genes$`Higher methylation in hEGCLCs_p10` %in% Genes_df$ensembl_gene_id_version]
```

```{r}
table(genes$`Higher methylation in hiPSCs`%in% Genes_df$ensembl_gene_id_version)
```

```{r}
DMG_uphiPSCsvshEGCLCs_p10 <- Genes_df[Genes_df$ensembl_gene_id_version %in% genes$`Higher methylation in hiPSCs`, "entrez_gene_id"]
DMG_uphiPSCsvshEGCLCs_p10 <- na.omit(DMG_uphiPSCsvshEGCLCs_p10[!DMG_uphiPSCsvshEGCLCs_p10 %in% ""])
DMG_downhiPSCsvshEGCLCs_p10 <- Genes_df[Genes_df$ensembl_gene_id_version %in% genes$`Higher methylation in hEGCLCs_p10`, "entrez_gene_id"]
DMG_downhiPSCsvshEGCLCs_p10 <- na.omit(DMG_downhiPSCsvshEGCLCs_p10[!DMG_downhiPSCsvshEGCLCs_p10 %in% ""])
```

```{r}
any(duplicated(DMG_uphiPSCsvshEGCLCs_p10))
```

```{r}
any(duplicated(DMG_downhiPSCsvshEGCLCs_p10))
```

```{r}
genes$`Higher methylation in hiPSCs` <- DMG_uphiPSCsvshEGCLCs_p10
genes$`Higher methylation in hEGCLCs_p10` <- DMG_downhiPSCsvshEGCLCs_p10
```

```{r, collapse=TRUE}
compKEGG <- compareCluster(geneCluster   = genes,
                         fun           = "enrichKEGG",
                         organism="hsa",
                         pvalueCutoff  = 0.05,
                         pAdjustMethod = "BH", 
                         universe = GeneUniverse_entrezid)
```

There are no significant terms (They have Padj>0.05)
```{r, fig.height=10, fig.width=15}
#dotplot(compKEGG, showCategory = 15, title = "KEGG Pathway Enrichment Analysis")
```

```{r, collapse=TRUE}
compGO <- compareCluster(geneCluster   = genes,
                         fun           = "enrichGO",
                         keyType = "ENTREZID",
                         pvalueCutoff  = 0.05,
                         pAdjustMethod = "BH", 
                         OrgDb='org.Hs.eg.db', 
                         universe=GeneUniverse_entrezid,
                         ont = "ALL")
```

```{r, fig.height=10, fig.width=15}
#dotplot(compGO, showCategory = 15, title = "GO Enrichment Analysis")
```

A new `DMRs_annotated` object is created in which annotation from `ChIPseeker` is added.

```{r, collapse=TRUE}
DMRs_annotated <- DMRs_GRanges

mcols(DMRs_annotated$hEGCLCs_p10vshiPSCs)$ChipSeekerAnn <- NA
mcols(DMRs_annotated$hEGCLCs_p10vshiPSCs)$ChipSeekerDistanceToTss <- NA
mcols(DMRs_annotated$hEGCLCs_p10vshiPSCs)$ChipSeekerGeneId <- NA

m <- findOverlaps(DMRs_annotated$hEGCLCs_p10vshiPSCs, peakAnnoList[["Higher methylation in hiPSCs"]]@anno)
mcols(DMRs_annotated$hEGCLCs_p10vshiPSCs)[queryHits(m), "ChipSeekerAnn"] <- mcols(peakAnnoList[["Higher methylation in hiPSCs"]]@anno)[subjectHits(m), "annotation"]
mcols(DMRs_annotated$hEGCLCs_p10vshiPSCs)[queryHits(m), "ChipSeekerDistanceToTss"] <- mcols(peakAnnoList[["Higher methylation in hiPSCs"]]@anno)[subjectHits(m), "distanceToTSS"]
mcols(DMRs_annotated$hEGCLCs_p10vshiPSCs)[queryHits(m), "ChipSeekerGeneId"] <- mcols(peakAnnoList[["Higher methylation in hiPSCs"]]@anno)[subjectHits(m), "geneId"]

m <- findOverlaps(DMRs_annotated$hEGCLCs_p10vshiPSCs, peakAnnoList[["Higher methylation in hEGCLCs_p10"]]@anno)
mcols(DMRs_annotated$hEGCLCs_p10vshiPSCs)[queryHits(m), "ChipSeekerAnn"] <- mcols(peakAnnoList[["Higher methylation in hEGCLCs_p10"]]@anno)[subjectHits(m), "annotation"]
mcols(DMRs_annotated$hEGCLCs_p10vshiPSCs)[queryHits(m), "ChipSeekerDistanceToTss"] <- mcols(peakAnnoList[["Higher methylation in hEGCLCs_p10"]]@anno)[subjectHits(m), "distanceToTSS"]
mcols(DMRs_annotated$hEGCLCs_p10vshiPSCs)[queryHits(m), "ChipSeekerGeneId"] <- mcols(peakAnnoList[["Higher methylation in hEGCLCs_p10"]]@anno)[subjectHits(m), "geneId"]
```

Adding entrez gene id, gene symbol and gene information

```{r}
DMRs_annotated$hEGCLCs_p10vshiPSCs <- as.data.frame(DMRs_annotated$hEGCLCs_p10vshiPSCs)
DMRs_annotated$hEGCLCs_p10vshiPSCs <- dplyr::left_join(DMRs_annotated$hEGCLCs_p10vshiPSCs, Genes_df, by=c('ChipSeekerGeneId' = 'ensembl_gene_id_version'))
```

```{r}
DMRs_annotated$hEGCLCs_p10vshiPSCs[!is.na(DMRs_annotated$hEGCLCs_p10vshiPSCs$hgnc_symbol),] %>% select(-c("ChipSeekerDistanceToTss", "ChipSeekerGeneId", "ensembl_gene_id", "entrez_gene_id")) %>% distinct() %>%
    datatable(class = 'hover', rownames = FALSE, caption="DMRs - hEGCLCs_p10vshiPSCs", extension='Buttons', escape = FALSE,
              options = list(pageLength=20, dom='Bfrtip', autoWidth=TRUE,
                             buttons=list(c('csv', 'excel'))))
```


```{r}
saveRDS(DMRs_annotated, paste0(OutputFolder, '/', 'DMRsAnnotated1.rds'))
```



## 4.2  iMeLCs VS hEGCLC_p10

### 4.2.1 Genetic annotation and visualization with __`ChIPseeker`__

DMRs are now annotated to their associated genes and genomic regions using the `annotatePeaks` function of the `ChIPseeker `R package, using a TxDb object. The [`GenomicFeatures`](https://bioconductor.org/packages/devel/bioc/vignettes/GenomicFeatures/inst/doc/GenomicFeatures.html) package uses `TxDb` objects to store transcript metadata. This class maps the 5’ and 3’ untranslated regions (UTRs), protein coding sequences (CDSs) and exons for a set of mRNA transcripts to their associated genome. 
All `TxDb` objects are backed by a SQLite database that manages genomic locations and the relationships between pre-processed mRNA transcripts, exons, protein coding sequences, and their related gene identifiers.

**TxDb created by us**

```{r, collapse=TRUE}
peakAnnoList <- lapply(list("Higher methylation in iMeLCs" = DMRs_GRanges$hEGCLCs_p10vsiMeLCs[DMRs_GRanges$hEGCLCs_p10vsiMeLCs$diff.Methy<0,], "Higher methylation in hEGCLCs_p10" = DMRs_GRanges$hEGCLCs_p10vsiMeLCs[DMRs_GRanges$hEGCLCs_p10vsiMeLCs$diff.Methy>0,]), annotatePeak, TxDb=txdb_v35, tssRegion=c(-3000, 3000), verbose=FALSE)

```

```{r, collapse=TRUE}
ChIPseeker::plotAnnoBar(peakAnnoList)
```

```{r, collapse=TRUE}
ChIPseeker::plotDistToTSS(peakAnnoList)
```

Given a list of gene set, `compareCluster` function will compute profiles of each gene cluster. 

```{r, collapse=TRUE}
genes = lapply(peakAnnoList, function(i) unique(as.data.frame(i)$geneId))
names(genes) = sub("_", "\n", names(genes))
```

```{r}
any(duplicated(genes$`Higher methylation in hEGCLC_p10`))
```

```{r}
any(duplicated(genes$`Higher methylation in iMeLCs`))
```

There are no duplicated ensembl ids

```{r}
table(genes$`Higher methylation in iMeLCs` %in% Genes_df$ensembl_gene_id_version)
```

```{r}
genes$`Higher methylation in iMeLCs` [!genes$`Higher methylation in iMeLCs` %in% Genes_df$ensembl_gene_id_version]
```

```{r}
table(genes$`Higher methylation in hEGCLC_p10`%in% Genes_df$ensembl_gene_id_version)
```

```{r}
DMG_uphEGCLCs_p10vsiMeLCs <- Genes_df[Genes_df$ensembl_gene_id_version %in% genes$`Higher methylation in hEGCLC_p10`, "entrez_gene_id"]
DMG_uphEGCLCs_p10vsiMeLCs <- na.omit(DMG_uphEGCLCs_p10vsiMeLCs[!DMG_uphEGCLCs_p10vsiMeLCs %in% ""])
DMG_downhEGCLCs_p10vsiMeLCs <- Genes_df[Genes_df$ensembl_gene_id_version %in% genes$`Higher methylation in iMeLCs`, "entrez_gene_id"]
DMG_downhEGCLCs_p10vsiMeLCs <- na.omit(DMG_downhEGCLCs_p10vsiMeLCs[!DMG_downhEGCLCs_p10vsiMeLCs %in% ""])
```

```{r}
any(duplicated(DMG_uphEGCLCs_p10vsiMeLCs))
```

```{r}
any(duplicated(DMG_downhEGCLCs_p10vsiMeLCs))
```

```{r}
genes$`Higher methylation in hEGCLC_p10` <- DMG_uphEGCLCs_p10vsiMeLCs
genes$`Higher methylation in iMeLCs` <- DMG_downhEGCLCs_p10vsiMeLCs
```

```{r, collapse=TRUE}
compKEGG <- compareCluster(geneCluster   = genes,
                         fun           = "enrichKEGG",
                         organism="hsa",
                         pvalueCutoff  = 0.05,
                         pAdjustMethod = "BH", 
                         universe = GeneUniverse_entrezid)
```

There are no significant terms (They have Padj>0.05)

```{r, fig.height=10, fig.width=15}
if (!is.null(compKEGG)) {
  dotplot(compKEGG, showCategory = 15, title = "KEGG Pathway Enrichment Analysis")
  }
```

```{r, collapse=TRUE}
compGO <- compareCluster(geneCluster   = genes,
                         fun           = "enrichGO",
                         keyType = "ENTREZID",
                         pvalueCutoff  = 0.05,
                         pAdjustMethod = "BH", 
                         OrgDb='org.Hs.eg.db', 
                         universe=GeneUniverse_entrezid,
                         ont = "ALL")
```

```{r, fig.height=10, fig.width=15}
#dotplot(compGO, showCategory = 15, title = "GO Enrichment Analysis")
```

A new `DMRs_annotated` object is created in which annotation from `ChIPseeker` is added.

```{r, collapse=TRUE}
DMRs_annotated <- DMRs_GRanges

mcols(DMRs_annotated$hEGCLCs_p10vsiMeLCs)$ChipSeekerAnn <- NA
mcols(DMRs_annotated$hEGCLCs_p10vsiMeLCs)$ChipSeekerDistanceToTss <- NA
mcols(DMRs_annotated$hEGCLCs_p10vsiMeLCs)$ChipSeekerGeneId <- NA


m <- findOverlaps(DMRs_annotated$hEGCLCs_p10vsiMeLCs, peakAnnoList[["Higher methylation in iMeLCs"]]@anno)
mcols(DMRs_annotated$hEGCLCs_p10vsiMeLCs)[queryHits(m), "ChipSeekerAnn"] <- mcols(peakAnnoList[["Higher methylation in iMeLCs"]]@anno)[subjectHits(m), "annotation"]
mcols(DMRs_annotated$hEGCLCs_p10vsiMeLCs)[queryHits(m), "ChipSeekerDistanceToTss"] <- mcols(peakAnnoList[["Higher methylation in iMeLCs"]]@anno)[subjectHits(m), "distanceToTSS"]
mcols(DMRs_annotated$hEGCLCs_p10vsiMeLCs)[queryHits(m), "ChipSeekerGeneId"] <- mcols(peakAnnoList[["Higher methylation in iMeLCs"]]@anno)[subjectHits(m), "geneId"]

m <- findOverlaps(DMRs_annotated$hEGCLCs_p10vsiMeLCs, peakAnnoList[["Higher methylation in hEGCLCs_p10"]]@anno)
mcols(DMRs_annotated$hEGCLCs_p10vsiMeLCs)[queryHits(m), "ChipSeekerAnn"] <- mcols(peakAnnoList[["Higher methylation in hEGCLCs_p10"]]@anno)[subjectHits(m), "annotation"]
mcols(DMRs_annotated$hEGCLCs_p10vsiMeLCs)[queryHits(m), "ChipSeekerDistanceToTss"] <- mcols(peakAnnoList[["Higher methylation in hEGCLCs_p10"]]@anno)[subjectHits(m), "distanceToTSS"]
mcols(DMRs_annotated$hEGCLCs_p10vsiMeLCs)[queryHits(m), "ChipSeekerGeneId"] <- mcols(peakAnnoList[["Higher methylation in hEGCLCs_p10"]]@anno)[subjectHits(m), "geneId"]


```

Adding entrez gene id, gene symbol and gene information

```{r}
DMRs_annotated$hEGCLCs_p10vsiMeLCs <- as.data.frame(DMRs_annotated$hEGCLCs_p10vsiMeLCs)
DMRs_annotated$hEGCLCs_p10vsiMeLCs <- dplyr::left_join(DMRs_annotated$hEGCLCs_p10vsiMeLCs, Genes_df, by=c('ChipSeekerGeneId' = 'ensembl_gene_id_version'))
```


```{r}
DMRs_annotated$hEGCLCs_p10vsiMeLCs[!is.na(DMRs_annotated$hEGCLCs_p10vsiMeLCs$hgnc_symbol),] %>% select(-c("ChipSeekerDistanceToTss", "ChipSeekerGeneId", "ensembl_gene_id", "entrez_gene_id")) %>% distinct() %>%
    datatable(class = 'hover', rownames = FALSE, caption="DMRs - iMeLCsvshEGCLC_p10", extension='Buttons', escape = FALSE,
              options = list(pageLength=20, dom='Bfrtip', autoWidth=TRUE,
                             buttons=list(c('csv', 'excel'))))
```



```{r}
saveRDS(DMRs_annotated, paste0(OutputFolder, '/', 'DMRsAnnotated2.rds'))
```




## 4.3 hPGCLCs vs hEGCLCs_p10

### 4.3.1 Genetic annotation and visualization with __`ChIPseeker`__

DMRs are now annotated to their associated genes and genomic regions using the `annotatePeaks` function of the `ChIPseeker `R package, using a TxDb object. The [`GenomicFeatures`](https://bioconductor.org/packages/devel/bioc/vignettes/GenomicFeatures/inst/doc/GenomicFeatures.html) package uses `TxDb` objects to store transcript metadata. This class maps the 5’ and 3’ untranslated regions (UTRs), protein coding sequences (CDSs) and exons for a set of mRNA transcripts to their associated genome. 
All `TxDb` objects are backed by a SQLite database that manages genomic locations and the relationships between pre-processed mRNA transcripts, exons, protein coding sequences, and their related gene identifiers.

**TxDb created by us**

```{r, collapse=TRUE}
peakAnnoList <- lapply(list("Higher methylation in hPGCLCs" = DMRs_GRanges$hEGCLCs_p10vshPGCLCs[DMRs_GRanges$hEGCLCs_p10vshPGCLCs$diff.Methy<0,], "Higher methylation in hEGCLC_p10" = DMRs_GRanges$hEGCLCs_p10vshPGCLCs[DMRs_GRanges$hEGCLCs_p10vshPGCLCs$diff.Methy>0,]), annotatePeak, TxDb=txdb_v35, tssRegion=c(-3000, 3000), verbose=FALSE)
```

```{r, collapse=TRUE}
ChIPseeker::plotAnnoBar(peakAnnoList)
```

```{r, collapse=TRUE}
ChIPseeker::plotDistToTSS(peakAnnoList)
```

Given a list of gene set, `compareCluster` function will compute profiles of each gene cluster. 

```{r, collapse=TRUE}
genes = lapply(peakAnnoList, function(i) unique(as.data.frame(i)$geneId))
names(genes) = sub("_", "\n", names(genes))
```

```{r}
any(duplicated(genes$`Higher methylation in hPGCLCs`))
```

```{r}
any(duplicated(genes$`Higher methylation in hEGCLC_p10`))
```

There are no duplicated ensembl ids

```{r}
table(genes$`Higher methylation in hEGCLC_p10` %in% Genes_df$ensembl_gene_id_version)
```

```{r}
genes$`Higher methylation in hEGCLC_p10` [!genes$`Higher methylation in hEGCLC_p10` %in% Genes_df$ensembl_gene_id_version]
```

```{r}
table(genes$`Higher methylation in hPGCLCs`%in% Genes_df$ensembl_gene_id_version)
```

```{r}
DMG_uphPGCLCsvshEGCLC_p10 <- Genes_df[Genes_df$ensembl_gene_id_version %in% genes$`Higher methylation in hPGCLCs`, "entrez_gene_id"]
DMG_uphPGCLCsvshEGCLC_p10 <- na.omit(DMG_uphPGCLCsvshEGCLC_p10[!DMG_uphPGCLCsvshEGCLC_p10 %in% ""])
DMG_downhPGCLCsvshEGCLC_p10 <- Genes_df[Genes_df$ensembl_gene_id_version %in% genes$`Higher methylation in hEGCLC_p10`, "entrez_gene_id"]
DMG_downhPGCLCsvshEGCLC_p10 <- na.omit(DMG_downhPGCLCsvshEGCLC_p10[!DMG_downhPGCLCsvshEGCLC_p10 %in% ""])
```

```{r}
any(duplicated(DMG_uphPGCLCsvshEGCLC_p10))
```

```{r}
any(duplicated(DMG_downhPGCLCsvshEGCLC_p10))
```

```{r}
genes$`Higher methylation in hPGCLCs` <- DMG_uphPGCLCsvshEGCLC_p10
genes$`Higher methylation in hEGCLC_p10` <- DMG_downhPGCLCsvshEGCLC_p10
```

```{r, collapse=TRUE}
compKEGG <- compareCluster(geneCluster   = genes,
                         fun           = "enrichKEGG",
                         organism="hsa",
                         pvalueCutoff  = 0.05,
                         pAdjustMethod = "BH", 
                         universe = GeneUniverse_entrezid)
```

There are no significant terms (They have Padj>0.05)
```{r, fig.height=10, fig.width=15}
#dotplot(compKEGG, showCategory = 15, title = "KEGG Pathway Enrichment Analysis")
```

```{r, collapse=TRUE}
compGO <- compareCluster(geneCluster   = genes,
                         fun           = "enrichGO",
                         keyType = "ENTREZID",
                         pvalueCutoff  = 0.05,
                         pAdjustMethod = "BH", 
                         OrgDb='org.Hs.eg.db', 
                         universe=GeneUniverse_entrezid,
                         ont = "ALL")
```

```{r, fig.height=10, fig.width=15}
dotplot(compGO, showCategory = 15, title = "GO Enrichment Analysis")
```

A new `DMRs_annotated` object is created in which annotation from `ChIPseeker` is added.

```{r, collapse=TRUE}
DMRs_annotated <- DMRs_GRanges

mcols(DMRs_annotated$hEGCLCs_p10vshPGCLCs)$ChipSeekerAnn <- NA
mcols(DMRs_annotated$hEGCLCs_p10vshPGCLCs)$ChipSeekerDistanceToTss <- NA
mcols(DMRs_annotated$hEGCLCs_p10vshPGCLCs)$ChipSeekerGeneId <- NA

m <- findOverlaps(DMRs_annotated$hEGCLCs_p10vshPGCLCs, peakAnnoList[["Higher methylation in hPGCLCs"]]@anno)
mcols(DMRs_annotated$hEGCLCs_p10vshPGCLCs)[queryHits(m), "ChipSeekerAnn"] <- mcols(peakAnnoList[["Higher methylation in hPGCLCs"]]@anno)[subjectHits(m), "annotation"]
mcols(DMRs_annotated$hEGCLCs_p10vshPGCLCs)[queryHits(m), "ChipSeekerDistanceToTss"] <- mcols(peakAnnoList[["Higher methylation in hPGCLCs"]]@anno)[subjectHits(m), "distanceToTSS"]
mcols(DMRs_annotated$hEGCLCs_p10vshPGCLCs)[queryHits(m), "ChipSeekerGeneId"] <- mcols(peakAnnoList[["Higher methylation in hPGCLCs"]]@anno)[subjectHits(m), "geneId"]

m <- findOverlaps(DMRs_annotated$hEGCLCs_p10vshPGCLCs, peakAnnoList[["Higher methylation in hEGCLC_p10"]]@anno)
mcols(DMRs_annotated$hEGCLCs_p10vshPGCLCs)[queryHits(m), "ChipSeekerAnn"] <- mcols(peakAnnoList[["Higher methylation in hEGCLC_p10"]]@anno)[subjectHits(m), "annotation"]
mcols(DMRs_annotated$hEGCLCs_p10vshPGCLCs)[queryHits(m), "ChipSeekerDistanceToTss"] <- mcols(peakAnnoList[["Higher methylation in hEGCLC_p10"]]@anno)[subjectHits(m), "distanceToTSS"]
mcols(DMRs_annotated$hEGCLCs_p10vshPGCLCs)[queryHits(m), "ChipSeekerGeneId"] <- mcols(peakAnnoList[["Higher methylation in hEGCLC_p10"]]@anno)[subjectHits(m), "geneId"]
```

Adding entrez gene id, gene symbol and gene information

```{r}
DMRs_annotated$hEGCLCs_p10vshPGCLCs <- as.data.frame(DMRs_annotated$hEGCLCs_p10vshPGCLCs)
DMRs_annotated$hEGCLCs_p10vshPGCLCs <- dplyr::left_join(DMRs_annotated$hEGCLCs_p10vshPGCLCs, Genes_df, by=c('ChipSeekerGeneId' = 'ensembl_gene_id_version'))
```

```{r}
DMRs_annotated$hEGCLCs_p10vshPGCLCs[!is.na(DMRs_annotated$hEGCLCs_p10vshPGCLCs$hgnc_symbol),] %>% select(-c("ChipSeekerDistanceToTss", "ChipSeekerGeneId", "ensembl_gene_id", "entrez_gene_id")) %>% distinct() %>%
    datatable(class = 'hover', rownames = FALSE, caption="DMRs - hEGCLCs_p10vshPGCLCs", extension='Buttons', escape = FALSE,
              options = list(pageLength=20, dom='Bfrtip', autoWidth=TRUE,
                             buttons=list(c('csv', 'excel'))))
```



```{r}
saveRDS(DMRs_annotated, paste0(OutputFolder, '/', 'DMRsAnnotated3.rds'))
```




## 4.4 hEGCLCs_p5 vs hEGCLCs_p10

### 4.4.1 Genetic annotation and visualization with __`ChIPseeker`__

DMRs are now annotated to their associated genes and genomic regions using the `annotatePeaks` function of the `ChIPseeker `R package, using a TxDb object. The [`GenomicFeatures`](https://bioconductor.org/packages/devel/bioc/vignettes/GenomicFeatures/inst/doc/GenomicFeatures.html) package uses `TxDb` objects to store transcript metadata. This class maps the 5’ and 3’ untranslated regions (UTRs), protein coding sequences (CDSs) and exons for a set of mRNA transcripts to their associated genome. 
All `TxDb` objects are backed by a SQLite database that manages genomic locations and the relationships between pre-processed mRNA transcripts, exons, protein coding sequences, and their related gene identifiers.

**TxDb created by us**


```{r, collapse=TRUE}
peakAnnoList <- lapply(list("Higher methylation in hEGCLCs_p5" = DMRs_GRanges$hEGCLCs_p10vshEGCLCs_p5[DMRs_GRanges$hEGCLCs_p10vshEGCLCs_p5$diff.Methy<0,], "Higher methylation in hEGCLCs_p10" = DMRs_GRanges$hEGCLCs_p10vshEGCLCs_p5[DMRs_GRanges$hEGCLCs_p10vshEGCLCs_p5$diff.Methy>0,]), annotatePeak, TxDb=txdb_v35, tssRegion=c(-3000, 3000), verbose=FALSE)
```

```{r, collapse=TRUE}
ChIPseeker::plotAnnoBar(peakAnnoList)
```

```{r, collapse=TRUE}
ChIPseeker::plotDistToTSS(peakAnnoList)
```

Given a list of gene set, `compareCluster` function will compute profiles of each gene cluster. 

```{r, collapse=TRUE}
genes = lapply(peakAnnoList, function(i) unique(as.data.frame(i)$geneId))
names(genes) = sub("_", "\n", names(genes))
```

```{r}
any(duplicated(genes$`Higher methylation in hEGCLCs_p5`))
```

```{r}
any(duplicated(genes$`Higher methylation in hEGCLCs_p10`))
```

There are no duplicated ensembl ids

```{r}
table(genes$`Higher methylation in hEGCLCs_p10` %in% Genes_df$ensembl_gene_id_version)
```

```{r}
genes$`Higher methylation in hEGCLCs_p10` [!genes$`Higher methylation in hEGCLCs_p10` %in% Genes_df$ensembl_gene_id_version]
```

```{r}
table(genes$`Higher methylation in hEGCLCs_p5`%in% Genes_df$ensembl_gene_id_version)
```

```{r}
DMG_uphEGCLCs_p5vshEGCLCs_p10 <- Genes_df[Genes_df$ensembl_gene_id_version %in% genes$`Higher methylation in hEGCLCs_p5`, "entrez_gene_id"]
DMG_uphEGCLCs_p5vshEGCLCs_p10 <- na.omit(DMG_uphEGCLCs_p5vshEGCLCs_p10[!DMG_uphEGCLCs_p5vshEGCLCs_p10 %in% ""])
DMG_downhEGCLCs_p5vshEGCLCs_p10 <- Genes_df[Genes_df$ensembl_gene_id_version %in% genes$`Higher methylation in hEGCLCs_p10`, "entrez_gene_id"]
DMG_downhEGCLCs_p5vshEGCLCs_p10 <- na.omit(DMG_downhEGCLCs_p5vshEGCLCs_p10[!DMG_downhEGCLCs_p5vshEGCLCs_p10 %in% ""])
```

```{r}
any(duplicated(DMG_uphEGCLCs_p5vshEGCLCs_p10))
```

```{r}
any(duplicated(DMG_downhEGCLCs_p5vshEGCLCs_p10))
```

```{r}
genes$`Higher methylation in hEGCLCs_p5` <- DMG_uphEGCLCs_p5vshEGCLCs_p10
genes$`Higher methylation in hEGCLCs_p10` <- DMG_downhEGCLCs_p5vshEGCLCs_p10
```

```{r, collapse=TRUE}
compKEGG <- compareCluster(geneCluster   = genes,
                         fun           = "enrichKEGG",
                         organism="hsa",
                         pvalueCutoff  = 0.05,
                         pAdjustMethod = "BH", 
                         universe = GeneUniverse_entrezid)
```

There are no significant terms (They have Padj>0.05)
```{r, fig.height=10, fig.width=15}
#dotplot(compKEGG, showCategory = 15, title = "KEGG Pathway Enrichment Analysis")
```

```{r, collapse=TRUE}
compGO <- compareCluster(geneCluster   = genes,
                         fun           = "enrichGO",
                         keyType = "ENTREZID",
                         pvalueCutoff  = 0.05,
                         pAdjustMethod = "BH", 
                         OrgDb='org.Hs.eg.db', 
                         universe=GeneUniverse_entrezid,
                         ont = "ALL")
```

```{r, fig.height=10, fig.width=15}
#dotplot(compGO, showCategory = 15, title = "GO Enrichment Analysis")
```

A new `DMRs_annotated` object is created in which annotation from `ChIPseeker` is added.

```{r, collapse=TRUE}
DMRs_annotated <- DMRs_GRanges

mcols(DMRs_annotated$hEGCLCs_p10vshEGCLCs_p5)$ChipSeekerAnn <- NA
mcols(DMRs_annotated$hEGCLCs_p10vshEGCLCs_p5)$ChipSeekerDistanceToTss <- NA
mcols(DMRs_annotated$hEGCLCs_p10vshEGCLCs_p5)$ChipSeekerGeneId <- NA

m <- findOverlaps(DMRs_annotated$hEGCLCs_p10vshEGCLCs_p5, peakAnnoList[["Higher methylation in hEGCLCs_p5"]]@anno)
mcols(DMRs_annotated$hEGCLCs_p10vshEGCLCs_p5)[queryHits(m), "ChipSeekerAnn"] <- mcols(peakAnnoList[["Higher methylation in hEGCLCs_p5"]]@anno)[subjectHits(m), "annotation"]
mcols(DMRs_annotated$hEGCLCs_p10vshEGCLCs_p5)[queryHits(m), "ChipSeekerDistanceToTss"] <- mcols(peakAnnoList[["Higher methylation in hEGCLCs_p5"]]@anno)[subjectHits(m), "distanceToTSS"]
mcols(DMRs_annotated$hEGCLCs_p10vshEGCLCs_p5)[queryHits(m), "ChipSeekerGeneId"] <- mcols(peakAnnoList[["Higher methylation in hEGCLCs_p5"]]@anno)[subjectHits(m), "geneId"]

m <- findOverlaps(DMRs_annotated$hEGCLCs_p10vshEGCLCs_p5, peakAnnoList[["Higher methylation in hEGCLCs_p10"]]@anno)
mcols(DMRs_annotated$hEGCLCs_p10vshEGCLCs_p5)[queryHits(m), "ChipSeekerAnn"] <- mcols(peakAnnoList[["Higher methylation in hEGCLCs_p10"]]@anno)[subjectHits(m), "annotation"]
mcols(DMRs_annotated$hEGCLCs_p10vshEGCLCs_p5)[queryHits(m), "ChipSeekerDistanceToTss"] <- mcols(peakAnnoList[["Higher methylation in hEGCLCs_p10"]]@anno)[subjectHits(m), "distanceToTSS"]
mcols(DMRs_annotated$hEGCLCs_p10vshEGCLCs_p5)[queryHits(m), "ChipSeekerGeneId"] <- mcols(peakAnnoList[["Higher methylation in hEGCLCs_p10"]]@anno)[subjectHits(m), "geneId"]
```

Adding entrez gene id, gene symbol and gene information

```{r}
DMRs_annotated$hEGCLCs_p10vshEGCLCs_p5 <- as.data.frame(DMRs_annotated$hEGCLCs_p10vshEGCLCs_p5)
DMRs_annotated$hEGCLCs_p10vshEGCLCs_p5 <- dplyr::left_join(DMRs_annotated$hEGCLCs_p10vshEGCLCs_p5, Genes_df, by=c('ChipSeekerGeneId' = 'ensembl_gene_id_version'))
```

```{r}
DMRs_annotated$hEGCLCs_p10vshEGCLCs_p5[!is.na(DMRs_annotated$hEGCLCs_p10vshEGCLCs_p5$hgnc_symbol),] %>% select(-c("ChipSeekerDistanceToTss", "ChipSeekerGeneId", "ensembl_gene_id", "entrez_gene_id")) %>% distinct() %>%
    datatable(class = 'hover', rownames = FALSE, caption="DMRs - hEGCLCs_p10vshEGCLCs_p5", extension='Buttons', escape = FALSE,
              options = list(pageLength=20, dom='Bfrtip', autoWidth=TRUE,
                             buttons=list(c('csv', 'excel'))))
```



```{r}
saveRDS(DMRs_annotated, paste0(OutputFolder, '/', 'DMRsAnnotated4.rds'))
```



# 5. Saving DMRs with annotation and Session Info

```{r SaveRDS}
saveRDS(Genes_df, paste0(OutputFolder, '/', 'GeneUniverse.rds'))
saveRDS(DMRs_annotated, paste0(OutputFolder, '/', 'DMRsAnnotated_final.rds'))
```

```{r SaveSession}
SessionInfo <- sessionInfo()
Date <- date()
```

```{r SessionInfo}
Date
SessionInfo
```