---
title: "Exploratory Data Analysis of Hsu et al WGBS"
author: "Gaja Matassa"
date: 'Date: `r format(Sys.Date(), "%B %d, %Y")`'
output:
    html_document:
        toc: true
        toc_float: true
        theme: 'united'
        highlight: 'kate'
        code_folding: hide
params: 
    InputFolder: '~/DataDir/7.EscapeesExploration/WGBS_Hsu'
    OutputFolder: '~/DataDir/7.EscapeesExploration/WGBS_Hsu/Output/'
---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE)

library(bsseq)
library(ggplot2)
library(ggrepel)
library(dplyr)
library(tidyr)
library(data.table)
library(DT)
library(dmrseq)
library(methylSig)

source("~/Transgenerational/EMSeq/NewData_fromLaSitua/Methylation_helper.R")
```

```{r, collapse=TRUE}
for (i in 1:length(params))
  print(paste('Parameter:', names(params)[i], ' - Value:', params[[i]], '- Class:', class(params[[i]])))
```

```{r, collapse=TRUE}
InputFolder <- params$InputFolder
OutputFolder <- params$OutputFolder

if (dir.exists(OutputFolder) == FALSE) {
  dir.create(OutputFolder, recursive=TRUE)
}
```

**Colors for figures**

```{r}
# sample_colors <- c("3_EG2_SL" = "#ff7f00", "3_EG9_SL" = "#ff7f00", "3_EG31_SL" = "#ff7f00",
#                    "2S" = "#dd1c77", "3S" = "#dd1c77", "5S" = "#dd1c77", "4S" = "#dd1c77",
#                    "2M" = "#377eb8", "3M" = "#377eb8", "5M" = "#377eb8", "4M" = "#377eb8",
#                    "2P" = "#4daf4a", "3P" = "#4daf4a", "5P" = "#4daf4a")
# 
cellTypes_colors <- c(hESC_KO ="#dd1c77", D4hPGCLC ="#4daf4a", hESC = "#ff7f00")
```

# 1. Loading data

BedGraph files in input are already merged (from individual Cytosines in a CpG to per-CpG metrics) and filtered (threshold: 10 reads). Here bedIntersect was already used to keep only on-bait CpGs.

```{r, collapse=TRUE}
bdg_files <-  list.files(path = InputFolder, pattern = '*sorted.bedGraph', full.names = TRUE) #output from our preprocessing
```

Creation of metadata from file names

```{r, collapse=TRUE}
sample_anno <- data.frame("SampleGEO" = sapply(strsplit(basename(bdg_files), split = "-|_"), function(x) x[1]),
           "Line" = c(sapply(strsplit(basename(bdg_files), split = "-|_")[1:10], function(x) x[13]), sapply(strsplit(basename(bdg_files), split = "-|_")[11:13], function(x) x[2])), 
           "TET1KO" = c(rep(FALSE,10), rep(TRUE,3)),
           "Type" = c(sapply(strsplit(basename(bdg_files), split = "-|_")[1:10], function(x) x[15]), 
                          paste0(sapply(strsplit(basename(bdg_files), split = "-|_")[11:13], function(x) x[5]), "_KO")),
           "Replicate" = c(sapply(strsplit(basename(bdg_files), split = "-|_")[1:2], function(x) x[17]), 
                           sapply(strsplit(basename(bdg_files), split = "-|_")[3:4], function(x) x[16]),
                           sapply(strsplit(basename(bdg_files), split = "-|_")[5:7], function(x) x[17]),
                           sapply(strsplit(basename(bdg_files), split = "-|_")[8:10], function(x) x[16]), 
                           sapply(strsplit(basename(bdg_files), split = "-|_")[11:13], function(x) x[4])),
           #"Sequencing" = rep("WGBS", length(bdg_files)),
           #"Day" = c(sapply(strsplit(basename(bdg_files), split = "-|_")[1:10], function(x) x[14]), NA, NA, NA)
           "FileName" = basename(bdg_files))
sample_anno$SampleID <- paste(sample_anno$Line, sample_anno$Type, sample_anno$Replicate, sep = "_")

if(!any(duplicated(sample_anno$SampleID))){
  rownames(sample_anno) <- sample_anno$SampleID
}else{stop("Duplicated sample names!")}
```

Match order

```{r, collapse=TRUE}
bdg_files<-bdg_files[order(match(basename(bdg_files), sample_anno$FileName))]
```

# 2. Reading bedgraph files with `read.bismark` from **bsseq** package

## 2.1 Generation of BSseq object

`strandCollapse = FALSE` because MethylDackel has an option to destrand data when methylation calls are made so that the output is already destranded.

```{r}
if(identical(basename(bdg_files), sample_anno$FileName)){
  bsseq_obj = bsseq::read.bismark(
      files = bdg_files,
      colData = sample_anno,
      rmZeroCov = FALSE,
      strandCollapse = FALSE #false if MethylDackel is used because MethylDackel has an option to destrand data when methylation calls are made so that the output is already destranded.
  )  
} else{stop("Sample names don't match between Bedgraph file and metadata")}
```

-   The number of samples in the BSSeq object is `r dim(bsseq_obj)[2]`.
-   The number of CpGs in the BSSeq object is `r dim(bsseq_obj)[1]`.

## 2.2 Exploration of BSseq object

```{r}
bsseq::pData(bsseq_obj)
```

`bsseq` package assumes that the following data has been extracted from alignments:

1.  genomic positions, including chromosome and location, for methylation loci.The genomic positions are stored in the bsseq object as a `GRanges` object. `GRanges` are general genomic regions; they represent a single base methylation loci as an interval of width 1 (which may seem a bit strange, but they said there are good reasons for this).

```{r}
head(GenomicRanges::granges(bsseq_obj))
```

**Methylation calls are from these chromosomes:**

```{r}
levels(bsseq_obj@rowRanges@seqnames@values)
```

2.  a (matrix) of M (Methylation) values, describing the number of read supporting methylation covering a single loci. Each row in this matrix is a methylation loci and each column is a sample.

```{r}
Meth_matrix <- bsseq::getCoverage(bsseq_obj, type = 'M') 
head(Meth_matrix)
```

3.  a (matrix) of Cov (Coverage) values, describing the total number of reads covering a single loci. Each row in this matrix is a methylation loci and each column is a sample.

```{r}
Cov_matrix <- bsseq::getCoverage(bsseq_obj, type = 'Cov')
head(Cov_matrix)
```

0/0 gives NaN, this is the case when 0 methylated reads mapping on that region over 0 total reads mapping on that region.

Methylation Proportions/Levels = Meth_matrix/Cov_matrix

```{r}
MethPerc_matrix <- (Meth_matrix/Cov_matrix)*100
head(MethPerc_matrix)
```

This matrix can also be obtained like this:

```{r}
head(getMeth(BSseq = bsseq_obj, type = "raw"))
```

**Check whether there are regions with NaNs for all samples**

Question: Is it true that there are no regions uncovered by any samples?

```{r}
all(apply(X = is.na(MethPerc_matrix), MARGIN = 1, FUN = sum) < ncol(MethPerc_matrix))
```

**Number of CpGs**

```{r, collapse=TRUE}
sample_anno$NumberCpGsPerSample <- colSums(Cov_matrix!=0)
sample_anno$NumberCpGsinBSseq <- rep(nrow(bsseq_obj), ncol(bsseq_obj))
sample_anno$PercCpGPerSampleinBSeq <- round((1-(colSums(Cov_matrix==0)/nrow(bsseq_obj)))*100, digits = 2)

datatable(sample_anno, extensions = "Buttons", 
            options = list(paging = TRUE,
                           scrollX=TRUE, 
                           searching = TRUE,
                           ordering = TRUE,
                           dom = 'Bfrtip',
                           buttons = c('copy', 'csv', 'excel', 'pdf'),
                           pageLength=10, 
                           lengthMenu=c(3,5,10) ))
```

# 3. Exploratory plots

The number of CpGs covered by at least one sample in the BSSeq object is `r dim(bsseq_obj)[1]`.

<!-- ## 3.1 Coverage distribution for all samples -->

<!-- ```{r, fig.width=13} -->
<!-- plotEmpiricalDistribution(bsseq_obj,  -->
<!--                           bySample = TRUE, -->
<!--                           testCovariate = NULL, -->
<!--                           type = "Cov") + guides(linetype="none") + ggplot2::scale_color_manual(values = sample_colors) -->
<!-- ``` -->

<!-- ```{r, fig.width=13} -->
<!-- plotEmpiricalDistribution(bsseq_obj,  -->
<!--                           bySample = TRUE, -->
<!--                           testCovariate = "Type", -->
<!--                           type = "Cov") + guides(linetype="none") + ggplot2::scale_color_manual(values = cellTypes_colors) -->
<!-- ``` -->

<!-- ## 3.2 Methylation levels distribution for all samples -->

<!-- ```{r, fig.width=13} -->
<!-- plotEmpiricalDistribution(bsseq_obj,  -->
<!--                           bySample = TRUE, -->
<!--                           testCovariate = NULL, -->
<!--                           adj = 3) + guides(linetype="none") #+ ggplot2::scale_color_manual(values = sample_colors) -->
<!-- ``` -->

<!-- ```{r, fig.width=13} -->
<!-- plotEmpiricalDistribution(bsseq_obj,  -->
<!--                           bySample = TRUE, -->
<!--                           testCovariate = "Type", -->
<!--                           adj = 3) + guides(linetype="none") #+ ggplot2::scale_color_manual(values = cellTypes_colors) -->
<!-- ``` -->

## 3.3 Violin Plot showing mean

```{r, collapse=TRUE, fig.width=10}
violin_plot(MethPerc_matrix, stat="mean", subset = 3000000) #+ ggplot2::scale_fill_manual(values = sample_colors)
```

## 3.4 Violin Plot showing median

```{r, collapse=TRUE, fig.width=10}
violin_plot(MethPerc_matrix, subset = 3000000) #+ ggplot2::scale_fill_manual(values = sample_colors)
```

**Mean values for Coverage**

```{r}
apply(Cov_matrix, 2, mean, na.rm = TRUE)
```

**Median values for Coverage**

```{r}
apply(Cov_matrix, 2, median, na.rm = TRUE)
```

**Mean values for Methylation Proportions**

```{r}
apply(MethPerc_matrix, 2, mean, na.rm = TRUE)
```

**Median values for Methylation Proportions**

```{r}
apply(MethPerc_matrix, 2, median, na.rm = TRUE)
```

## 3.5 PCA plots

```{r, collapse=TRUE}
pca_res <- do_PCA(MethPerc_matrix)
```

```{r, collapse=TRUE, fig.width=8}
plot_PCA(pca_res = pca_res, anno = sample_anno, col_anno = "Type", shape_anno = "Line", custom_colors = cellTypes_colors, point_size = 5)
```

<!-- ## 3.6 Dendogram plot -->

<!-- ```{r, fig.height=4, fig.width=8} -->
<!-- d <- hclust(dist(t(MethPerc_matrix), -->
<!--                  method="euclidean")) -->
<!-- #d$labels <- rownames(pData(bsseq_obj)) -->
<!-- d$labels <- rownames(sample_anno) -->
<!--   #paste0(pData(bsseq_obj)$Type, "_", pData(bsseq_obj)$InputDNA) -->
<!-- dend <- as.dendrogram(d) -->
<!-- plot(dend) -->
<!-- ``` -->

# 4. Filtering BSSeq object by shared CpGs

```{r}
table(sample_anno$Type)
dim(methylSig::filter_loci_by_group_coverage(
    bs = bsseq_obj,
    group_column = 'Type',
    min_samples_per_group = c('D4hPGCLC' = 5, 'hESC' = 5, 'hESC_KO' = 3)))[1]
```

```{r}
#length(which(apply(X = is.na(getMeth(bsseq_obj, type="raw")), MARGIN = 1, FUN = sum) == 0)) #26830070
```

-   The number of samples in the BSSeq object is `r dim(bsseq_obj)[2]`.
-   The number of CpGs in the BSSeq object is `r dim(bsseq_obj)[1]`.
-   The number of CpGs covered by at least 50% of samples in the BSSeq object is `r length(which(apply(X = is.na(getMeth(bsseq_obj, type="raw")), MARGIN = 1, FUN = sum) <= round(0.5*dim(bsseq_obj)[2], digits=0)))`.
-   The number of CpGs covered by at least 75% of samples in the BSSeq object is `r length(which(apply(X = is.na(getMeth(bsseq_obj, type="raw")), MARGIN = 1, FUN = sum) <= round(0.25*dim(bsseq_obj)[2], digits=0)))`.
-   The number of CpGs covered by all samples in the BSSeq object is `r length(which(apply(X = is.na(getMeth(bsseq_obj, type="raw")), MARGIN = 1, FUN = sum) == 0))`.
-   The number of CpGs covered by at least 60% of samples of the same type in the BSSeq object is `r dim(methylSig::filter_loci_by_group_coverage(bs = bsseq_obj,group_column = "Type", min_samples_per_group = c('D4hPGCLC' = 3, 'hESC' = 3, 'hESC_KO' = 2)))[1]`.

# 5. Saving and Session Info

Saving BSSeq objects :

-   Complete

```{r SavingI, collapse=TRUE}
saveRDS(bsseq_obj, file = paste0 (OutputFolder, "bsseq_obj_complete.rds"))
```

-   Filtered by sample and by CpGs covered by all samples kept

```{r SavingII, collapse=TRUE}
pData(bsseq_obj)$Seq <- "WGBS"
saveRDS(methylSig::filter_loci_by_group_coverage(bs = bsseq_obj,group_column = "Seq", min_samples_per_group = c("WGBS" = dim(bsseq_obj)[2])), file = paste0 (OutputFolder, "bsseq_obj_sharedbyall.rds")) #same of doing group_column = 'Type', min_samples_per_group = c('D4hPGCLC' = 5, 'hESC' = 5, 'hESC_KO' = 3))
```

-   With CpGs covered by 75% of all samples kept (independently on cell types)

```{r SavingIII, collapse=TRUE}
saveRDS(methylSig::filter_loci_by_group_coverage(bs = bsseq_obj,group_column = "Seq", min_samples_per_group = c("WGBS" = round(0.75*dim(bsseq_obj)[2], digits=0))), file = paste0 (OutputFolder, "bsseq_obj_sharedby75ofall.rds"))
```

- With CpGs covered by  at least 60% of samples of the same cell type

```{r SavingIV, collapse=TRUE}
saveRDS(methylSig::filter_loci_by_group_coverage(bs = bsseq_obj,group_column = "Type", min_samples_per_group = c('D4hPGCLC' = 3, 'hESC' = 3, 'hESC_KO' = 2)), file = paste0 (OutputFolder, "bsseq_obj_sharedby60CellType.rds"))
```

Session Info

```{r SaveSession}
SessionInfo <- sessionInfo()
Date <- date()
```

```{r SessionInfo}
Date
SessionInfo
```
