---
title: "Exploratory Data Analysis of Hsu et al"
author: "Gaja Matassa"
date: 'Date: `r format(Sys.Date(), "%B %d, %Y")`'
output:
    html_document:
        toc: true
        toc_float: true
        theme: 'united'
        highlight: 'kate'
        code_folding: hide
params: 
    InputFolder_WGBS: '~/DataDir/7.EscapeesExploration/WGBS_Hsu/Output/'
    InputFolder_bACEseq: '~/DataDir/7.EscapeesExploration/bACEseq_Hsu/Output/'
    OutputFolder: ''
---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE)

library(bsseq)
library(ggplot2)
library(ggrepel)
library(dplyr)
library(tidyr)
library(data.table)
library(DT)
library(dmrseq)
library(methylSig)
library(GenomicRanges)
source("~/Transgenerational/EMSeq/NewData_fromLaSitua/Methylation_helper.R")
```

```{r, collapse=TRUE}
for (i in 1:length(params))
  print(paste('Parameter:', names(params)[i], ' - Value:', params[[i]], '- Class:', class(params[[i]])))
```

```{r, collapse=TRUE}
InputFolder_WGBS <- params$InputFolder_WGBS
InputFolder_bACEseq <- params$InputFolder_bACEseq

# if (dir.exists(OutputFolder) == FALSE) {
#   dir.create(OutputFolder, recursive=TRUE)
# }
```

From [Hsu et al. paper](10.1016/j.isci.2023.107191):
_A major function of TET1 is the oxidation of 5mC to 5hmC,39 where it can serve as a relatively stable epigenetic mark in non-dividing cells, or as an intermediate for active DNA demethylation. To examine 5hmC enrichment in hPGCLCs, we profiled genome-wide 5hmC levels at single-base resolution using bisulfite-assisted ACEseq (bACEseq).29 By pairing bACEseq with whole genome bisulfite sequencing (WGBS), true 5mC levels and sites can be inferred by subtracting bACEseq signals (5hmC only) from WGBS signals (5mC + 5hmC mixed readout), thereby resolving the epigenetic base ambiguity between 5mC and 5hmC. By profiling ∼40 million (M) common CpG sites between UCLA1 and UCLA2 hESCs and hPGCLCs, we identified a doubling of 5hmC levels in hPGCLCs coupled with a modest reduction of 5mC that can be partially explained by the presence of 5hmC (Figure 2A)._

Therefore I decide for our downstream exploration to create a **bsseq object with genomic loci shared by the two techniques WGBS and bACEseq and shared by UCLA1 and UCLA2 hESCs and hPGCLCs. It will have assays with methylation levels of 5mCs and 5hmCs separated.**

```{r}
bsseq_obj_WGBS <- readRDS(paste0(InputFolder_WGBS, "bsseq_obj_complete.rds"))
bsseq_obj <- readRDS(paste0(InputFolder_bACEseq, "bsseq_obj_complete.rds"))
```

```{r}
table(pData(bsseq_obj)$Type)
table(pData(bsseq_obj_WGBS)$Type)
bsseq_obj_WGBS <- methylSig::filter_loci_by_group_coverage(
    bs = bsseq_obj_WGBS,
    group_column = 'Type',
    min_samples_per_group = c('D4hPGCLC' = 5, 'hESC' = 5, 'hESC_KO' = 0))
bsseq_obj <- methylSig::filter_loci_by_group_coverage(
    bs = bsseq_obj,
    group_column = 'Type',
    min_samples_per_group = c('D4hPGCLC' = 6, 'hESC' = 4, 'hESC_KO' = 0))
```


```{r}
#uniqueInbACEseq <- GenomicRanges::setdiff(bsseq_obj, bsseq_obj_WGBS)
hits <- findOverlaps(granges(bsseq_obj_WGBS), granges(bsseq_obj), type = "equal")
bsseq_WGBS_common <- bsseq_obj_WGBS[queryHits(hits), ]
bsseq_bACE_common <- bsseq_obj[subjectHits(hits), ]
```

Between the two techniques it doesn't seem that we have the exact same samples. In particular, for UCLA1 hESCs only one replicate is shared and for UCLA2 hESCs there is one more replicate for WGBS than for bACEseq. Furthermore, for UCLA1 hPGCLCs there is one more replicate for bACEseq than WGBS. How to deal with it?
Exclude the additional UCLA1 hPGCLC replicate and make a fake pairment of UCLA1_hESC_rep2 in WGBS and UCLA1_hESC_rep1 in bACEseq? Or just keep only the shared replicate for UCLA1 hESCs?

```{r}
rownames(pData(bsseq_WGBS_common))
rownames(pData(bsseq_bACE_common))
```

```{r}
samples_to_remove <- c("UCLA1_hESC_rep1", "UCLA1_hESC_rep2", "UCLA2_hESC_rep1", "UCLA1_D4hPGCLC_rep1", "UCLA1_hESC_KO_CDKO1102", "UCLA1_hESC_KO_CDKO1105", "UCLA1_hESC_KO_CDKO1312")
bsseq_WGBS <- bsseq_WGBS_common[ , !rownames(pData(bsseq_WGBS_common)) %in% samples_to_remove]
rm(bsseq_WGBS_common)
bsseq_bACE <- bsseq_bACE_common[ , !rownames(pData(bsseq_bACE_common)) %in% samples_to_remove]
rm(bsseq_bACE_common)
```

```{r}
rownames(pData(bsseq_WGBS))
rownames(pData(bsseq_bACE))
```

Keeping only samples that are shared by both the techniques (without considering the 3 KO hESCs) we go from 10 samples to 8. Let's proceed with them

```{r}
granges(bsseq_WGBS)
granges(bsseq_bACE)
```
```{r}
if(!identical(granges(bsseq_WGBS), granges(bsseq_bACE))){
  common_seqs <- intersect(seqlevels(bsseq_obj_WGBS), seqlevels(bsseq_obj))
  seqlevels(bsseq_WGBS, pruning.mode="coarse") <- common_seqs
  seqlevels(bsseq_bACE, pruning.mode="coarse") <- common_seqs}
rm(bsseq_obj_WGBS)
rm(bsseq_obj)
```

```{r}
if (identical(rownames(pData(bsseq_WGBS)), rownames(pData(bsseq_bACE))) & identical(granges(bsseq_WGBS), granges(bsseq_bACE))) {
  bsseq_complete <- bsseq_WGBS
  names(assays(bsseq_complete)) <- c("M_WGBS", "Cov_WGBS")
  assays(bsseq_complete)$Meth_WGBS <- getMeth(BSseq = bsseq_WGBS, type = "raw")
  names(assays(bsseq_complete)) <- c("M_WGBS", "Cov_WGBS", "Meth_WGBS")
  assays(bsseq_complete)$M_bACEseq <- getCoverage(bsseq_bACE, type = 'M') 
  assays(bsseq_complete)$Cov_bACEseq <- getCoverage(bsseq_bACE, type = 'Cov') 
  assays(bsseq_complete)$Meth_bACEseq <- getMeth(BSseq = bsseq_bACE, type = "raw")
}
```

```{r}
assays(bsseq_complete)$Meth_hmC <- assays(bsseq_complete)$Meth_bACEseq
assays(bsseq_complete)$Meth_mC <- assays(bsseq_complete)$Meth_WGBS - assays(bsseq_complete)$Meth_bACEseq

summary(assays(bsseq_complete)$Meth_mC)

#cap values between 0 and 1 (to avoid small negatives)
assays(bsseq_complete)$Meth_mC[assays(bsseq_complete)$Meth_mC < 0] <- 0
assays(bsseq_complete)$Meth_mC[assays(bsseq_complete)$Meth_mC > 1] <- 1

summary(assays(bsseq_complete)$Meth_mC)
```

Violin plot showing global trend (and mean) of 5mCs + 5hmCs
```{r}
violin_plot(assays(bsseq_complete)$Meth_WGBS, stat="mean", subset = 3000000)
```

Violin plot showing global trend (and mean) of only 5hmCs
```{r}
violin_plot(assays(bsseq_complete)$Meth_hmC, stat="mean", subset = 3000000)
```

Violin plot showing global trend (and mean) of only 5mCs
```{r}
violin_plot(assays(bsseq_complete)$Meth_mC, stat="mean", subset = 3000000)
```

```{r}
saveRDS(bsseq_complete, "~/DataDir/7.EscapeesExploration/bsseq_Hsu_complete.rds")
saveRDS(bsseq_WGBS, "~/DataDir/7.EscapeesExploration/bsseq_WGBS.rds")
saveRDS(bsseq_bACE, "~/DataDir/7.EscapeesExploration/bsseq_bACE.rds")
```

