---
title: 'Comparison escapees vs non-escapees'
author: 'Gaja Matassa'
date: 'Date: `r format(Sys.Date(), "%B %d, %Y")`'
output:
    html_document:
        toc: true
        toc_float: true
        theme: 'united'
        highlight: 'kate'
        code_folding: hide
params: 
    InputFolder: '~/DataDir/7.EscapeesExploration/Output/EscapeeBsseq/'
    num_permutations: 500
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, warning = FALSE)
```

```{r}
library(plyr)
library(dplyr)
library(bsseq)
library(dmrseq)
library(ggplot2)
source("../Methylation_helper.R")
```

Set random seed and number of permutations
```{r}
set.seed(123)
num_permutations <- params$num_permutations
```

# 1. Data loading and fast exploration by PCA

Loading bsseq objects

```{r}
bsseq_obj <- readRDS("~/DataDir/1.PreliminaryAnalysis/Output/WithSelectedEGCLCs/bsseq_obj_sharedby75ofall.rds") #getting bsseq object where sample selection and CpG filtering were already performed 
bsseq_obj_escapees <- readRDS(paste0(params$InputFolder, "bsseq_obj_escapees.rds"))
bsseq_obj_NonEscapees <- readRDS(paste0(params$InputFolder, "bsseq_obj_NonEscapees.rds"))
```

* Our bsseq object contain `r dim(bsseq_obj)[1]` CpG loci.
* Among them `r dim(bsseq_obj_escapees)[1]` CpG loci belong to escapees regions (escapee CpGs).
* While `r dim(bsseq_obj_NonEscapees)[1]` CpG loci do not belong to escapees regions (non-escapee CpGs).

**Colors for figures**

```{r}
cellTypes_colors <- c(hiPSCs ="#dd1c77", iMeLCs ="#377eb8", hPGCLCs ="#4daf4a", hEGCLCs = "#ff7f00")
```

## PCA only on escapee CpGs

```{r, collapse=TRUE, fig.width=8}
pca_res <- do_PCA(getMeth(bsseq_obj_escapees, type = "raw"))
sample_anno <- bsseq::pData(bsseq_obj_escapees)
plot_PCA(pca_res = pca_res, anno = sample_anno, col_anno = "Type", shape_anno = NULL, custom_colors = cellTypes_colors, point_size = 5)
```

## PCA only on non-escapee CpGs

```{r, collapse=TRUE, fig.width=8}
pca_res <- do_PCA(getMeth(bsseq_obj_NonEscapees, type = "raw"))
sample_anno <- bsseq::pData(bsseq_obj_NonEscapees)
plot_PCA(pca_res = pca_res, anno = sample_anno, col_anno = "Type", shape_anno = NULL, custom_colors = cellTypes_colors, point_size = 5)
```

As expected it resembles the PCA done on the complete one, since we removed only `r dim(bsseq_obj_escapees)[1]` out of a total of `r dim(bsseq_obj)[1]` CpG loci.

# 2.Generation of `r num_permutations` random sets of `r dim(bsseq_obj_escapees)[1]` non-escapee CpGs and distributions plots

Let's generate random indices to randomly select `r dim(bsseq_obj_escapees)[1]` CpG loci among the non-escapee ones.

```{r}
random_sets <- list()

for (i in 1:num_permutations) {
  random_indices <- sample(1:dim(bsseq_obj_NonEscapees)[1], dim(bsseq_obj_escapees)[1], replace = FALSE)
  bsseq_obj_randomNonEscapees <- bsseq_obj_NonEscapees[random_indices, ]
  
  random_sets[[i]] <- bsseq_obj_randomNonEscapees
}
```

**Head of BSSEQ with escapee CpGs**

```{r, collapse=FALSE}
getMeth(bsseq_obj_escapees, type = "raw") %>% head()
```

**Escapee CpG methylation distribution plot**

```{r}
plotEmpiricalDistribution(bsseq_obj_escapees, 
                          bySample = TRUE,
                          testCovariate = "Type",
                          adj = 3) + guides(linetype="none") + 
  labs(title = "Escapee CpG methylation distribution by cell type") + ggplot2::scale_color_manual(values = cellTypes_colors)
```

```{r}
plotEmpiricalDistribution(bsseq_obj_escapees, 
                          bySample = FALSE,
                          testCovariate = "Line",
                          adj = 3) + guides(linetype="none") + 
  labs(title = "Escapee CpG methylation distribution collapsing cell types")
```

**Non-Escapee CpG methylation distribution plot**

```{r}
plotEmpiricalDistribution(bsseq_obj_NonEscapees, 
                          bySample = TRUE,
                          testCovariate = "Type",
                          adj = 3) + guides(linetype="none") + 
  labs(title = "Non-escapee CpG methylation distribution by cell type") + ggplot2::scale_color_manual(values = cellTypes_colors)
```

```{r}
plotEmpiricalDistribution(bsseq_obj_NonEscapees, 
                          bySample = FALSE,
                          testCovariate = "Line",
                          adj = 3) + guides(linetype="none") + 
  labs(title = "Non-escapee CpG methylation distribution collapsing cell types")
```

**Head of first BSSEQ with randomly selected non-escapee CpGs**

```{r, collapse=FALSE}
getMeth(random_sets[[1]], type = "raw") %>% head()
```

**First set of randomly selected non-escapee CpG methylation distribution plot**

```{r}
plotEmpiricalDistribution(random_sets[[1]], 
                          bySample = TRUE,
                          testCovariate = "Type",
                          adj = 3) + guides(linetype="none") + 
  labs(title = "Non-escapee CpG methylation distribution by cell type (1)") + ggplot2::scale_color_manual(values = cellTypes_colors)
```

**Head of second BSSEQ with randomly selected non-escapee CpGs**

```{r, collapse=FALSE}
getMeth(random_sets[[2]], type = "raw") %>% head()
```

**Second set of randomly selected non-escapee CpG methylation distribution plot**

```{r}
plotEmpiricalDistribution(random_sets[[2]], 
                          bySample = TRUE,
                          testCovariate = "Type",
                          adj = 3) + guides(linetype="none") + 
  labs(title = "Non-escapee CpG methylation distribution by cell type (2)") + ggplot2::scale_color_manual(values = cellTypes_colors)
```

# 3. Kolmogorov-Smirnov test

Example of KS test result comparing distributions in case of hEGCLC_rep1 sample: escapee CpG methylation levels vs first randomly selected non-escapee set.

```{r, collapse=FALSE}
ks_test_result <- ks.test(getMeth(bsseq_obj_escapees, type = "raw")[, "hEGCLC_rep1"], getMeth(random_sets[[1]], type = "raw")[, "hEGCLC_rep1"])
print(ks_test_result)
```

Example of KS test result comparing distributions in case of hEGCLC_rep1 sample: first randomly selected non-escapee set vs second randomly selected non-escapee set.

```{r, collapse=FALSE}
ks_test_result <- ks.test(getMeth(random_sets[[1]], type = "raw")[, "hEGCLC_rep1"], getMeth(random_sets[[2]], type = "raw")[, "hEGCLC_rep1"])
print(ks_test_result)
```

Example of KS test result comparing distributions: escapee CpG methylation levels (hiPSC_rep2 sample) vs first randomly selected non-escapee set (hPGCLC_rep2 sample).

```{r, collapse=FALSE}
ks_test_result <- ks.test(getMeth(random_sets[[1]], type = "raw")[, "hiPSC_rep2"], getMeth(random_sets[[1]], type = "raw")[, "hPGCLC_rep2"])
print(ks_test_result)
```

Example of KS test result comparing distributions: escapee CpG methylation levels vs first randomly selected non-escapee set **considering the mean levels across cell types (rowwise)**.

```{r, collapse=FALSE}
ks_test_result <- ks.test(getMeth(bsseq_obj_escapees, type = "raw") %>% rowMeans(na.rm = TRUE), getMeth(random_sets[[1]], type = "raw") %>% rowMeans(na.rm = TRUE))
print(ks_test_result)
```

Example of KS test result comparing distributions: first randomly selected non-escapee set vs second randomly selected non-escapee set **considering the mean levels across cell types (rowwise)**.

```{r, collapse=FALSE}
ks_test_result <- ks.test(getMeth(random_sets[[1]], type = "raw") %>% rowMeans(na.rm = TRUE), getMeth(random_sets[[2]], type = "raw") %>% rowMeans(na.rm = TRUE))
print(ks_test_result)
```

*Let's calculate the mean methylation levels across cell types in all sets of randomly selected non-escapee CpGs.*

```{r}
random_MeanMethLevels <- lapply(random_sets, function(bsseq){
  getMeth(bsseq, type = "raw") %>% rowMeans(na.rm = TRUE)})
```

**Let's do KS test comparing escapee CpG methylation levels vs all `r num_permutations` randomly selected non-escapee sets.**
```{r, collapse=FALSE}
ks_test_results <- lapply(random_MeanMethLevels, function(random_MeanMeth) {
  ks.test(getMeth(bsseq_obj_escapees, type = "raw") %>% rowMeans(na.rm = TRUE), random_MeanMeth)
})
pvalues_kstest <- sapply(ks_test_results, function(result) {
  result$p.value
})
adjusted_pvalues_kstest <- p.adjust(pvalues_kstest, method = "fdr")
print(adjusted_pvalues_kstest)
```

How many adjusted p-values are < 0.05?

```{r, collapse=FALSE}
table(adjusted_pvalues_kstest < 0.05)
```

**Let's do KS test comparing first randomly selected non-escapee set vs all `r num_permutations` randomly selected non-escapee sets.**

```{r, collapse=FALSE}
ks_test_results <- lapply(random_MeanMethLevels, function(random_MeanMeth) {
  ks.test(random_MeanMethLevels[[1]], random_MeanMeth)
})
pvalues_kstest <- sapply(ks_test_results, function(result) {
  result$p.value
})
adjusted_pvalues_kstest <- p.adjust(pvalues_kstest, method = "fdr")
print(adjusted_pvalues_kstest)
```

How many adjusted p-values are < 0.05?

```{r, collapse=FALSE}
table(adjusted_pvalues_kstest < 0.05)
```

# 4.Escapee and non escapee CpG methylation distribution plot 

```{r}
MeanMethLevels_df <- c(unlist(random_MeanMethLevels), getMeth(bsseq_obj_escapees, type = "raw") %>% rowMeans(na.rm = TRUE) %>% as.vector()) %>% as.data.frame()
colnames(MeanMethLevels_df) <- "MeanMeth"
MeanMethLevels_df$Distribution <- c(paste0("Random_", rep(1:num_permutations, each = dim(bsseq_obj_escapees)[1])), rep("Escapees_CpGs", each = dim(bsseq_obj_escapees)[1]))
MeanMethLevels_df$Category <- "Non Escapees"
MeanMethLevels_df$Category[(nrow(MeanMethLevels_df)+1-nrow(bsseq_obj_escapees)):nrow(MeanMethLevels_df)] <- "Escapees"
```

```{r, fig.width=10}
ggplot(MeanMethLevels_df, aes(x = MeanMeth, color = Category, group = Distribution)) +
  geom_density(alpha = 0.5)  +
  labs(
    title = "Distribution CpG mean methylation levels in escapees and non-escapees regions", 
    x = "Mean Methylation Levels across cell types", 
    y = "Density", 
    color = "Category"
  ) +
  scale_color_manual(values = c("Escapees" = "#FF7D00", "Non Escapees" = "#15616D")) + 
  theme_minimal()
```

# 5.Session Info

```{r}
date()
sessionInfo()
```

