---
title: "Generation of bsseq object with escapee CpGs provided by MitinoriLab (Murase et al. paper)"
author: "Gaja Matassa"
date: 'Date: `r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_document:
    toc: yes
    toc_float: yes
    theme: united
    highlight: kate
    code_folding: hide
  word_document:
    toc: yes
  pdf_document:
    toc: yes
params:
  InputFolder: '~/DataDir/7.EscapeesExploration/Escapees_Murase/'
  OutputFolder: '~/DataDir/7.EscapeesExploration/Escapees_Murase/Output/'
---

# 1. Environment setting

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, warning = FALSE)
options(scipen = 100, digits = 4)
```

```{r, collapse=TRUE}
library(ggplot2)
library(bsseq)
library(scales)
library(gridExtra)
library(dplyr)
library(dmrseq)
library(annotatr)
library(sechm)
library(rtracklayer)
source("../Methylation_helper.R")
```

```{r, collapse=TRUE}
OutputFolder <- params$OutputFolder
```

# 2. Loading data

## 2.1 BSSeq object

Loading of bsseq object with CpGs shared by 75% of the samples.

```{r, collapse=TRUE}
bsseq_obj <- readRDS("~/DataDir/1.PreliminaryAnalysis/Output/WithSelectedEGCLCs/bsseq_obj_sharedby75ofall.rds") #getting bsseq object where sample selection and CpG filtering were already performed 
```

## 2.3 Escapees regions

These escapees regions were provided by Mitinori lab and are both from the [Tang et al. 2015](https://doi.org/10.1016/j.cell.2015.04.053) and [Murase et al.](https://doi.org/10.1038/s41586-024-07526-6) papers.

```{r}
library(data.table)

EscapeesRegions_F <- fread(
  "~/DataDir/7.EscapeesExploration/Escapees_Murase/hypermrF5_F1cF1csF2cF2aghGCF_hypermr_merge_sum_homer_label.bed",
  sep = "\t",
  header = FALSE,
  fill = TRUE  # fills in missing columns with NA
)

EscapeesRegions_M <- fread(
  "~/DataDir/7.EscapeesExploration/Escapees_Murase/hypermrM4_M1_M1s_M2_hGC_hypermr_merge_sum_homer_label.bed",
  sep = "\t",
  header = FALSE,
  fill = TRUE  # fills in missing columns with NA
)
```

```{r}
head(EscapeesRegions_M)
EscapeesRegions_M <- EscapeesRegions_M[, c(1,2,3,15,16,18,19,20)]
colnames(EscapeesRegions_M) <- c("Chr","Start","End","GenomicType","HomerAnno","Gene","Repetitive","Repetitive2")
```

Let's now focus on escapees found in the male line.
We have `r nrow(EscapeesRegions_M)` escapees regions.

**These regions are in GRCh38.p12 assembly and with 0-based coordinates**

```{r}
EscapeesRegions_GRanges <- makeGRangesFromDataFrame(EscapeesRegions_M, seqnames.field = "Chr", start.field = "Start", end.field = "End", keep.extra.columns = FALSE, na.rm=TRUE, starts.in.df.are.0based=TRUE)
```

# 3. Generation of BSSeq oject with escapee CpGs and non-escapee CpGs

```{r}
bsseq_ranges_df <- granges(bsseq_obj) %>% as.data.frame() #granges has 0-based position they seem 1-based. Look here for explanation: https://www.biostars.org/p/84686/
bsseq_ranges_df$end <- bsseq_ranges_df$start + 1 #to make them 0-based
bsseq_granges <- makeGRangesFromDataFrame(bsseq_ranges_df, starts.in.df.are.0based = TRUE)

bsseq_obj_1based <- bsseq_obj
bsseq_obj_1based@rowRanges <- bsseq_granges
granges(bsseq_obj_1based)
```

This to obtain a bsseq object with only CpGs contained in the escapees regions

```{r}
m <- findOverlaps(granges(bsseq_obj_1based), EscapeesRegions_GRanges)

bsseq_obj_escapees <- bsseq_obj_1based[queryHits(m),]
```

With our targeted kit we are able to cover **`r length(bsseq_obj_escapees)`** CpGs belonging to the escapees regions.

Let's create also a bsseq object with non-escapee CpGs (all the other CpGs that are not in escapees regions)

```{r}
bsseq_obj_NonEscapees <- bsseq_obj_1based[-queryHits(m),]
```

# 4. Violin plot showing methylation levels of escapee CpGs

```{r}
assays(bsseq_obj_escapees)$Meth <- getMeth(bsseq_obj_escapees, type = "raw")
assays(bsseq_obj_NonEscapees)$Meth <- getMeth(bsseq_obj_NonEscapees, type = "raw")

sample_colors <- c("hiPSC_rep1" = "#dd1c77", "hiPSC_rep2" = "#dd1c77", "hiPSC_rep3" = "#dd1c77", "hiPSC_rep4" = "#dd1c77",
                   "iMeLC_rep1" = "#377eb8", "iMeLC_rep2" = "#377eb8", "iMeLC_rep3" = "#377eb8", "iMeLC_rep4" = "#377eb8",
                   "hPGCLC_rep1" = "#4daf4a", "hPGCLC_rep2" = "#4daf4a", "hPGCLC_rep3" = "#4daf4a",
                   "hEGCLC_rep1" = "#ff7f00", "hEGCLC_rep2" = "#ff7f00", "hEGCLC_rep3" = "#ff7f00")

violin_plot(assays(bsseq_obj_escapees)$Meth, stat="median") + ggplot2::scale_fill_manual(values = sample_colors)
```

# 5. Saving and Session Info

Saving:

-   DataFrame of Escapee regions

-   bsseq object subset with only CpGs in escapees regions

-   bsseq object subset with non-escapee CpGs (all the other CpGs that are not in escapees regions)

```{r}
openxlsx::write.xlsx(data.frame(
  seqnames = as.character(seqnames(EscapeesRegions_GRanges)),
  start = start(EscapeesRegions_GRanges),
  end = end(EscapeesRegions_GRanges),
  strand = as.character(strand(EscapeesRegions_GRanges)),
  mcols(EscapeesRegions_GRanges)
), file = paste0(OutputFolder, "EscapeeRegions_Murase_hg38.xlsx")) 
saveRDS(bsseq_obj_escapees, paste0(OutputFolder, "bsseq_obj_escapees_Murase.rds"))
saveRDS(bsseq_obj_NonEscapees, paste0(OutputFolder, "bsseq_obj_NonEscapees_Murase.rds"))
```

```{r}
date()
sessionInfo()
```
